---
# Canonical source. Notes use Quarto/reveal.js Speaker View (S).
# No automatic title slide: the first explicit ## heading is slide 1.
# Default: native R output/plots. Optional: captured MCP output via output_source:mcp.
title: ""
pagetitle: "MCP Console · R, Python, and SQL together"
engine: knitr
params:
  output_source: r
format:
  revealjs:
    slide-level: 2
    output-file: mcp-console.html
    show-notes: false
    width: 1600
    height: 900
    margin: 0
    center: false
    transition: none
    slide-number: c/t
    embed-resources: true
    code-copy: false
    code-line-numbers: false
    controls: true
    progress: true
    css: styles.css
execute:
  eval: false
  echo: false
  warning: false
  message: false
---

```{r}
#| label: render-setup
#| eval: true
#| include: false
source("examples/render-helpers.R", local = knitr::knit_global())
```

## MCP Console {#interactive-work .deck-slide .hero .story-slide}
::: {.kicker}

Why Console exists

:::

::: {.big-statement}

R, Python, and SQL.<br>Available in one session.

:::

::: {.takeaway}

Let the model choose the language the task needs.

:::

::: {.notes}

**Say:** I want the model to have R, Python, and SQL available in the same session, and choose the language that helps with the task. That keeps R’s packages and methods in reach alongside Python’s ecosystem and SQL database queries.

We also keep solving the same execution problems around each language: waiting, output, dependencies, records, and permissions. Console puts that work into a shared runtime, with the sandbox as part of the design from the beginning.

**Show:** Start with why. The small tool interface is a consequence of this design, which we will see in a few concrete examples.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::

## Choose the language for the task {#small-surface .deck-slide .story-slide}
::: {.kicker}

Why Console exists

:::

:::: {.three-col}

::: {.story-card}

**R**

Statistical methods, analysis packages, and data visualization.

:::

::: {.story-card}

**Python**

Machine learning libraries and the broader Python ecosystem.

:::

::: {.story-card}

**SQL**

Local or remote database connections. Optimized queries.

:::

::::
::: {.takeaway}

The model can choose the language and syntax it works best with.

:::

::: {.notes}

**Say:** These are capabilities available to one model. R can fit the model and visualize the results. Python can use the same data with another library. SQL can query a local or remote database, letting its engine optimize filters, joins, and aggregations. The model can choose as the task develops, including the syntax it works best with. Data visualization is available in both R and Python; these cards do not assign exclusive strengths. A selected DBI or Python DB-API connection supplies the SQL backend; remote access needs the relevant driver, credentials, and network permission.

Keeping R useful here means making its capabilities easy to reach from the agent’s ordinary workflow. We should not need to choose an R-only or Python-only product before the work begins.

**Show:** Each language has reasons to use it. The examples are illustrative uses, not exclusive assignments of what a language can do.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::

## Solve the execution problems once {#shared-problems .deck-slide .story-slide}
::: {.kicker}

Why Console exists

:::

:::: {.three-col}

::: {.story-card}

**Keep work alive**

Persistent objects.<br>Wait, poll, and interrupt.<br>Interactive input.

:::

::: {.story-card}

**Make results usable**

Text and plots.<br>Compact progress updates.<br>Retained output and records.

:::

::: {.story-card}

**Control execution**

Managed dependencies.<br>Filesystem and network policy.<br>Local and remote hosts.

:::

::::
::: {.takeaway}

R, Python, and SQL share this infrastructure.

:::

::: {.notes}

**Say:** These are the problems we often solve piecemeal around a language tool. Duplicating the runtime means solving much of this again for each language. Console shares the execution lifecycle, output handling, recording, and sandbox boundary. Language-specific adapters still do their own work.

The aim is one coherent execution system. I’ll use a few small details later to show what that means in practice.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [Architecture](https://github.com/t-kalinowski/mcp-console/blob/main/docs/ARCHITECTURE.md)

:::

## Full language capabilities, with control over risk {#sandbox-motivation .deck-slide .story-slide}
::: {.kicker}

Why Console exists

:::

:::: {.three-col}

::: {.story-card}

**Use real tools**

The full languages, their libraries, native code, and subprocesses.

:::

::: {.story-card}

**Provide real guarantees**

Enforce file and network permissions. Own the worker’s process cleanup.

:::

::: {.story-card}

**Make it configurable**

Choose the files, services, and execution environment the task can use.

:::

::::

::: {.takeaway}

Safety belongs in the execution system, alongside the capabilities.

:::

::: {.notes}

**Say:** We want the model to have the languages’ full capabilities. That includes libraries, native code, and subprocesses. Those capabilities also carry risks: a mistake can change files, read secrets, contact a remote service, or leave work running.

The sandbox constrains what that code can access, and the runner owns its lifecycle. Safety depends on those permissions and on trusted setup. This does not make arbitrary execution risk-free. Package preparation runs outside the worker sandbox, and the default host-read policy needs explicit denials for confidential files.

**Show:** Connect full language capabilities to enforced permissions and cleanup, then to the user’s configuration choices. The default policy and cleanup behavior follow after the examples.

**Sources:** [Sandbox configuration](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md) · [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::

## Start with a cell {#send-r .deck-slide .literal-slide}
::: {.kicker}

One analysis · R and Python together

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

LLM client → Console

:::

<!-- cells:fit -->
```python
console.send(r="""
  d <- read.csv("examples/measurements.csv")
  fit <- lm(response ~ temperature + group,
            data = d)
  fit
""")
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

R output

:::

```{r}
#| label: fit-output
#| eval: true
#| echo: false
#| comment: ""
console_text("fit")
```

::::

:::::

::::::
::: {.takeaway}

Define objects in the session and see their printed output.

:::

::: {.notes}

**Say:** The final expression is fit, so R calls the fitted object’s default print method. The model receives that printed text; the fitted object remains in the R session. The agent receives useful output immediately, not prose that stands in for output. This is a real, persistent R workspace; ordinary cells do not require another object-management API.

**Show:** Show the exact submitted R code beside the printed lm object. The code and renderer both read the same cell definition. The data frame stays available for the following cells.

**Author / capture:** Default preview evaluates this cell in native R through knitr, not in MCP Console. With -P output_source:mcp, the panel reads the literal captured Console text. No simulated numerical output is committed.

**Sources:** [runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [send](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SEND_OPERATIONS.md)

:::

## Return plots {#r-plots .deck-slide .literal-slide}
::: {.kicker}

One analysis · R and Python together

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

LLM client → Console

:::

<!-- cells:r-plot -->
```python
console.send(r="""
  plot(d$temperature, d$response,
       xlab = "Temperature",
       ylab = "Response")
""")
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

R graphics output

:::

```{r}
#| label: r-plot-output
#| eval: true
#| echo: false
#| comment: ""
#| fig-width: 7
#| fig-height: 5.2
#| dpi: 144
#| dev: png
#| out-width: "100%"
console_plot("r-plot")
```

::::

:::::

::::::
::: {.takeaway}

R graphics return an image.

:::

::: {.notes}

**Say:** This is the base-R plot produced by this exact plot() call. Plotting is part of the runtime contract, not another tool. All drawing for a managed plot belongs in the same cell.

**Show:** Let R render this figure during quarto preview. Do not substitute a Matplotlib graphic. Capture mode uses the PNG returned by Console instead.

**Author / capture:** Default uses the displayed base-R cell through knitr; plot capture mode reads captures/r-plot-01.png. Neither a stock image nor the previous illustrative SVG is used.

**Sources:** [runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::

## Send Python to the same session {#python-first .deck-slide .literal-slide}
::: {.kicker}

One analysis · R and Python together

:::


:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

Python cell

:::

<!-- cells:python-first -->
```python
console.send(python='print(len(r.d))')
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

Console response

:::

```{r}
#| eval: true
#| echo: false
#| comment: ""
console_protocol("python-first")
```

::::

:::::

::::::

::: {.takeaway}

Supply `r=` or `python=`. Python can read the R data frame through `r.d`.

:::

::: {.notes}

**Say:** We have sent R code to load the data, fit a model, and draw a plot. The model can also send Python code through the same tool by choosing the python argument. Python works in the same session. Start with one expression: how many rows are in the R data frame? The r bridge exposes the existing object. Next we can use Python libraries on that data.

**Show:** This is an actual captured Python response.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::

## Use a Python library on the same data {#python .deck-slide .code-dense}
::: {.kicker}

One analysis · R and Python together

:::

:::: {.code-panel}

::: {.panel-label}

LLM client → Console · Python cell

:::

<!-- cells:python-ml -->
```python
console.send(python="""
from sklearn.ensemble import RandomForestRegressor
from sklearn.model_selection import cross_val_score
import pandas as pd

frame = r.d
X = pd.get_dummies(frame[["temperature", "group"]])
model = RandomForestRegressor(random_state=42)
scores = cross_val_score(model, X, frame["response"],
                         cv=5, scoring="neg_mean_absolute_error")
print(f"CV MAE: {-scores.mean():.3f}")
""")
```

::::


:::: {.result-panel}

::: {.panel-label}

Captured Python result

:::

```{r}
#| eval: true
#| echo: false
#| comment: ""
console_protocol("python-ml")
```

::::

::: {.notes}

**Say:** We already have the data in R. Python reads it through r.d and uses scikit-learn for cross-validation. This shows why both languages belong in one session: the choice follows the library we want to use. The numerical result is a real capture; comparing the statistical merits of the two models is outside this example.

**Show:** The meaningful difference is the library being used: RandomForestRegressor and cross_val_score. The result is shown below the call.

**Sources:** [runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [requirements](https://github.com/t-kalinowski/mcp-console/blob/main/docs/REQUIREMENTS.md)

:::

## Query the same data with SQL {#sql .deck-slide .literal-slide}
::: {.kicker}

One analysis · R and Python together

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

SQL cell

:::

<!-- cells:sql -->
```python
console.send(sql="""
SELECT "group", COUNT(*) AS n
FROM d
GROUP BY "group"
ORDER BY "group"
""")
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

Captured result

:::

```{r}
#| eval: true
#| echo: false
#| comment: ""
console_protocol("sql")
```

::::

:::::

::::::

::: {.notes}

**Say:** The same session can also query the live R data frame with DuckDB. The model can choose SQL for a grouped count without exporting the frame first. The connection and catalog persist across calls.

**Show:** d is the R data frame from the first example. This response comes from the same captured Console session.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::

## SQL uses the connection you choose {#sql-connections .deck-slide .literal-slide}
::: {.kicker}

One analysis · choose the database connection in R or Python

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

R · a DBI connection

:::

```python
console.send(r="""
  con <- DBI::dbConnect(
    RSQLite::SQLite(), ":memory:")
  console_sql_connection(con)
""")
```

::::

:::::

::::: {.right}

:::: {.code-panel}

::: {.panel-label}

Python · a DB-API connection

:::

```python
console.send(python="""
import sqlite3
con = sqlite3.connect(":memory:")
console_sql_connection(con)
""")
```

::::

:::::

::::::

:::: {.code-panel}

::: {.panel-label}

Subsequent SQL cells use the selected connection

:::

```python
console.send(sql="SELECT CURRENT_TIMESTAMP AS now")
```

::::

::: {.takeaway}

Use a local database or an existing remote connection through its driver.

:::

::: {.notes}

**Say:** SQL defaults to the session’s DuckDB connection. The model can instead select a valid DBI connection owned by R, or a DB-API connection owned by Python. These SQLite examples are alternatives; a remote database connection follows the same selection pattern. The driver supplies the SQL dialect and transaction state.

Later sql calls go to the selected connection. The object stays in its owning runtime, with its connection-local state. These are supported database connection interfaces, not arbitrary objects. A selected database sees its own tables and registrations; it does not automatically inherit the default DuckDB connection’s access to R data frames.

Restore the managed connection with console_sql_connection(NULL) in R or console_sql_connection(None) in Python before closing a custom connection. A remote connection still needs its driver, credentials, and network permissions.

**Show:** Read the R and Python panels as two alternatives, followed by the same SQL call. No model or remote database call is needed for this illustration.

**Sources:** [SQL connection selection](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md#sql)

:::

## Make CRAN and PyPI available to the session {#package-opportunity .deck-slide .story-slide}
::: {.kicker}

Use the package ecosystem

:::

::: {.story-lead}

A model may avoid a useful package to avoid changing the user’s environment.

:::

:::::: {.two-col}

::::: {.left}

::: {.story-card}

**Let the task choose the library**

The model can use a trusted package when the analysis calls for it.

:::

:::::

::::: {.right}

::: {.story-card}

**Manage the environment for the session**

Console retains requirements and prepares compatible package environments.

:::

:::::

::::::
::: {.takeaway}

Ordinary package use can request the dependency it needs.

:::

::: {.notes}

**Say:** In my experience, models often hesitate to install packages because installation sounds like a change to the user’s global setup. They may settle for a simpler approach using only what is already there. I want using the right library to be an ordinary part of the analysis.

Here the requirement belongs to the Console session and is resolved into a managed environment. That puts the breadth of trusted CRAN and PyPI packages in reach. The session’s retained requirement configuration is temporary; resolver caches and prepared environments may be reused on disk. This does not mean every package installs successfully or that all its installation code is sandboxed.

**Sources:** [Requirements and trust](https://github.com/t-kalinowski/mcp-console/blob/main/docs/REQUIREMENTS.md)

:::

## Separate session control from code execution {#process-topology .deck-slide .story-slide}
::: {.kicker}

How the processes fit together · local execution

:::

<!-- inline-svg:assets/process-topology.svg -->
```{=html}
<div class="embedded-diagram">
<svg xmlns="http://www.w3.org/2000/svg" viewBox="0 0 1400 455" role="img" aria-label="The LLM client talks to the Console server. The server uses trusted resolver processes outside the sandbox and exchanges cells and results with the sandboxed relay and worker. The native runner enforces permissions and cleanup.">
<g font-family="Arial, sans-serif" fill="#18323f">
<rect x="865" y="8" width="520" height="435" rx="18" fill="#edf5ef" stroke="#167d76" stroke-width="2" stroke-dasharray="8 6"/>
<text x="1125" y="43" text-anchor="middle" font-size="24" fill="#167d76">Sandboxed processes</text>
<rect x="10" y="83" width="245" height="100" rx="10" fill="white" stroke="#a9b6b6"/>
<text x="132" y="142" text-anchor="middle" font-size="29">LLM client</text>
<rect x="390" y="83" width="305" height="100" rx="10" fill="white" stroke="#a9b6b6"/>
<text x="542" y="125" text-anchor="middle" font-size="29">Console server</text>
<text x="542" y="158" text-anchor="middle" font-size="23">Session and records</text>
<path d="M265 133 H380" fill="none" stroke="#537b79" stroke-width="2"/>
<text x="320" y="108" text-anchor="middle" font-size="23">MCP</text>
<rect x="930" y="83" width="390" height="100" rx="10" fill="white" stroke="#a9b6b6"/>
<text x="1125" y="125" text-anchor="middle" font-size="29">Worker relay</text>
<text x="1125" y="158" text-anchor="middle" font-size="23">Transport and supervision</text>
<path d="M705 133 H920" fill="none" stroke="#537b79" stroke-width="2"/>
<text x="780" y="108" text-anchor="middle" font-size="22">Cells / results</text>
<rect x="930" y="290" width="390" height="105" rx="10" fill="#d9eee9" stroke="#167d76"/>
<text x="1125" y="330" text-anchor="middle" font-size="29">Session worker</text>
<text x="1125" y="367" text-anchor="middle" font-size="25">R · Python · SQL</text>
<path d="M1125 194 V278" fill="none" stroke="#537b79" stroke-width="2"/>
<text x="1144" y="239" font-size="23">Cells and I/O</text>
<rect x="390" y="290" width="305" height="105" rx="10" fill="#fff" stroke="#a9b6b6"/>
<text x="542" y="330" text-anchor="middle" font-size="29">Trusted resolvers</text>
<text x="542" y="367" text-anchor="middle" font-size="25">ir · uv</text>
<path d="M542 194 V278" fill="none" stroke="#537b79" stroke-width="2"/>
<text x="560" y="239" font-size="23">Prepare packages</text>
<text x="1125" y="428" text-anchor="middle" font-size="21" fill="#167d76">Native runner: permissions and cleanup</text>
</g>
<g fill="#537b79"><path d="M265 133 l10 -5 v10 Z M380 133 l-10 -5 v10 Z M705 133 l10 -5 v10 Z M920 133 l-10 -5 v10 Z M1125 194 l-5 10 h10 Z M1125 278 l-5 -10 h10 Z M542 194 l-5 10 h10 Z M542 278 l-5 -10 h10 Z"/></g>
</svg>
</div>
```
<!-- /inline-svg -->

::: {.takeaway}

The server manages the session. The worker executes code inside the sandbox.

:::

::: {.notes}

**Say:** The LLM client talks to the Console server over MCP. The server owns the session and its records. A relay carries cells, input, output, and control messages to the worker, where R, Python, and SQL run. The native runner applies the sandbox policy and owns process cleanup.

Package resolvers are separate trusted processes, outside the worker sandbox. That is how the server can prepare a package without giving the worker direct network access. We will follow one of those requests next.

**Show:** This is the simplified local layout. Arrows indicate the labeled two-way exchanges, not a strict parent-process tree. Native setup helpers are omitted. The native runner is a separate executable; the shaded area identifies the processes whose permissions it restricts. Remote targets change placement and transport while retaining the session interface.

**Sources:** [Implemented architecture](https://github.com/t-kalinowski/mcp-console/blob/main/docs/ARCHITECTURE.md) · [Sandbox configuration](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::

## Use the package; let the runtime prepare it {#r-resolution .deck-slide .story-slide .literal-slide}
::: {.kicker}

Use the package ecosystem

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

R

:::

```python
console.send(r="library(data.table)")
```

::::

:::::

::::: {.right}

:::: {.code-panel}

::: {.panel-label}

Python

:::

```python
console.send(python="import polars")
```

::::

:::::

::::::
::: {.sequence-strip}

Reach a missing package → resolve → activate → continue

:::

::: {.takeaway}

The live workspace stays available during supported additions.

:::

::: {.notes}

**Say:** The model writes ordinary R or Python. When execution reaches a supported missing-package operation, the runtime requests preparation, activates the result, and continues the original operation. It keeps the live workspace.

For R this uses ir; managed Python uses uv. We react to package use during execution rather than pre-scanning the source or replaying the whole cell. Explicit requirements remain available when the model needs a version pin or a distribution name that cannot be inferred from an import.

**Show:** These are illustrative package-use calls. Availability depends on a working managed resolver and the package’s system prerequisites. Docker’s prepared-image path is different and appears later.

**Sources:** [Requirements and trust](https://github.com/t-kalinowski/mcp-console/blob/main/docs/REQUIREMENTS.md)

:::

## Follow a package-resolution request {#resolver-boundary .deck-slide .story-slide .literal-slide}
::: {.kicker}

Private worker protocol · illustrative paths

:::

:::::: {.two-col}

::::: {.left}

::: {.resolution-steps}

1. R reaches a missing package and the worker requests resolution.
2. The server prepares a library with `ir`, outside the worker sandbox.
3. The worker activates the library and continues the original load.

:::

:::::

::::: {.right}

:::: {.code-panel}

::: {.panel-label}

Worker → server

:::

```json
{"kind": "resolve_r",
 "packages": ["data.table"]}
```

::::

:::: {.code-panel}

::: {.panel-label}

Server → worker

:::

```json
{"kind": "r_resolved",
 "library": "/cache/r/library"}
```

::::

:::: {.code-panel}

::: {.panel-label}

Worker → server

:::

```json
{"kind": "r_activated",
 "library": "/cache/r/library"}
```

::::

:::::

::::::

::: {.takeaway}

Trusted preparation can use the network while the worker remains restricted.

:::

::: {.notes}

**Say:** The request comes from the runtime inside the worker, through the relay to the server. The server validates it and uses a trusted resolver outside the sandbox to prepare the library. The result is a library path, which the worker adds to its live library search path. It acknowledges activation before continuing the package load. The server commits the retained environment only for a matching activation from the current worker generation.

These are private worker-protocol messages, not MCP tool calls the model has to make. The field names match the documented protocol; /cache/r/library is an illustrative path. Python follows the analogous resolve_python exchange and activates a prepared interpreter environment.

Installation and build code run with the preparation account’s permissions, so requirements and resolver configuration must be trusted. Caches can persist. The worker’s filesystem and network policy continues to apply. On SSH, preparation runs on the remote execution host.

**Sources:** [Worker protocol](https://github.com/t-kalinowski/mcp-console/blob/main/docs/WORKER_PROTOCOL.md#nested-managed-r-resolution) · [Requirements and trust](https://github.com/t-kalinowski/mcp-console/blob/main/docs/REQUIREMENTS.md)

:::

## Collect progress while the code keeps running {#wait-timeout .deck-slide .literal-slide .code-dense}
::: {.kicker}

Detail 1 · Waiting preserves the computation

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

LLM client → Console · running cell

:::

<!-- cells:progress -->
```python
console.send(r="""
  pb <- txtProgressBar(max = 20, style = 3, width = 20)
  for (i in 1:20) {
    Sys.sleep(0.15)
    setTxtProgressBar(pb, i)
  }
  close(pb)
""", timeout_ms=450)
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

Console → text · captured response

:::

```{r}
#| label: progress-first
#| eval: true
#| echo: false
#| comment: ""
console_protocol("progress")
```

::::
::: {.small-note}

The percentage depends on when the wait expires.

:::

:::::

::::::

::: {.notes}

**Say:** This uses R’s built-in txtProgressBar and setTxtProgressBar; no extra package is needed. Sys.sleep stands in for each unit of work. A calculation can take longer than a tool call should hold the client. Here the client waits 450 milliseconds and receives the progress so far. The computation is still running in the same worker. The timeout describes how long to wait for output. We can poll for more without submitting the calculation again.

**Show:** Read both the compact progress line and the exact running notice. The panel reads the captured response; call boundaries remain timing-dependent.

**Author / capture:** Warm up the session before capturing this example. Do not promise a particular percentage at 450 ms. The local capture script retains actual responses rather than asserting fixed percentages.

**Sources:** [send](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SEND_OPERATIONS.md) · [runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::

## Pick up the next output from the same run {#poll .deck-slide .literal-slide}
::: {.kicker}

Detail 1 · Waiting and readable progress

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

LLM client → Console · collect new output

:::

<!-- cells:poll-next,poll-final -->
```python
console.send(timeout_ms=600)

console.send(timeout_ms=5000)
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

Console → text · after 600 ms

:::

```{r}
#| label: poll-first
#| eval: true
#| echo: false
#| comment: ""
console_protocol("poll-next")
```

::::

:::: {.result-panel}

::: {.panel-label}

Console → text · after 5000 ms

:::

```{r}
#| label: poll-last
#| eval: true
#| echo: false
#| comment: ""
console_protocol("poll-final")
```

::::
:::::

::::::
::: {.takeaway}

Poll the same run. Redrawn progress stays one current line per response.

:::

::: {.notes}

**Say:** These are separate responses to later polls, not a transcript dumped into one response. Each poll contains newly collected output. Carriage-return redraws within one response interval collapse to its current line; a later interval can return the next current line.

**Show:** Point to the two poll calls and their corresponding results. The last response has real output and no invented [done] appended. Emphasize that empty send() is also a poll.

**Author / capture:** Both rendering modes read the two literal captured responses. The wire exchange records the 600 ms poll followed by the 5000 ms poll. Exact progress fractions are not a public API guarantee.

**Sources:** [runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [send](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SEND_OPERATIONS.md)

:::

## Keep the response readable; retain the longer output {#bounded-output .deck-slide .literal-slide .code-dense}
::: {.kicker}

Detail 2 · Preserve evidence beyond the context window

:::


:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

Call

:::

<!-- cells:flood -->
```python
console.send(r=r"""
  for (i in seq_len(10000)) {
    cat(sprintf("row %05d\n", i))
  }
  cat("fit complete\n")
""")
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

One tool response · shortened for display

:::

```text
row 00001
row 00002
row 00003

[output preview: middle omitted;
 raw output retained at
 .agents/console/sessions/<run>/
 outputs/call-000008.log]

row 09999
row 10000
fit complete
```

::::

:::::

::::::

::: {.notes}

**Say:** A noisy cell should not use the whole context window. One response keeps a beginning, a recent tail, and a notice pointing to retained output. The model can inspect the longer record with an appropriate filesystem tool if needed. The text budget is 8 KiB per result, with a separate allowance for images; raw retention also has a limit. This slide shortens the response for display.

**Show:** This is one stylized response. The middle notice and path are shortened, and most preview lines are omitted to fit the slide. The complete literal response is in captures/flood.txt.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::

## Capture output below the language hooks {#retained-output .deck-slide .literal-slide}
::: {.kicker}

Detail 2 · Native output belongs in the result

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

R · forked child writes through a pipe

:::

<!-- cells:r-fork-cat -->
```python
console.send(r=r"""
  job <- parallel::mcparallel({
    out <- pipe("cat", "w")
    cat("hello from fork\n", file = out)
    close(out)
  })
  invisible(parallel::mccollect(job))
""")
```

::::

:::: {.result-panel}

::: {.panel-label}

Console response

:::

```{r}
#| eval: true
#| echo: false
#| comment: ""
console_protocol("r-fork-cat")
```

::::

:::::

::::: {.right}

:::: {.code-panel}

::: {.panel-label}

Python · write directly to stdout

:::

<!-- cells:python-fd -->
```python
console.send(python=r"""
import os
os.write(1, b"hello directly on fd 1\n")
""")
```

::::

:::: {.result-panel}

::: {.panel-label}

Console response

:::

```{r}
#| eval: true
#| echo: false
#| comment: ""
console_protocol("python-fd")
```

::::

:::::

::::::

::: {.takeaway}

Capture native output from the worker and its children.

:::

::: {.notes}

**Say:** The R example forks with parallel::mcparallel. Inside that child, R’s cat writes to a pipe; the native cat command forwards the bytes to stdout. Closing the pipe finishes that command, and mccollect waits for the forked child. This uses Unix process APIs and was captured on macOS.

The Python example writes directly to file descriptor 1, bypassing sys.stdout. The text is followed by 23, the number of bytes returned by os.write and displayed as the cell’s final expression. There is no assignment to discard that value.

**Show:** Both panels use actual Console responses. Read each call with its own result. The R child → pipe → native cat route is deliberate: ordinary cat() in the fork completed but its text was absent in this installed build. Do not present the plain call as captured successfully. The older C-level fork example remains in the reference captures.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::

## The complete send() interface {#full-signature .deck-slide .signature-slide}
::: {.kicker}

Work with the session

:::

::: {.full-contract}

<pre class="signature-display">console.send(
  <span class="syntax-accent">r</span>=…, <span class="syntax-accent">python</span>=…, <span class="syntax-accent">sql</span>=…,  <span class="syntax-comment"># at most one language</span>
  control=…,               <span class="syntax-comment"># interrupt or restart</span>
  stdin=…,                 <span class="syntax-comment"># input for a read or prompt</span>
  requirements={r: …, python: …, duckdb: …},
  timeout_ms=…,            <span class="syntax-comment"># how long this call waits</span>
)</pre>

:::

::: {.takeaway}

Omit code to poll, supply input, control the session, or prepare dependencies.

:::

::: {.notes}

**Say:** These are the seven top-level fields in the managed session's send interface. All are optional, and a call accepts at most one complete R, Python, or SQL cell. Requirements contain lists of R packages, Python distributions, or managed DuckDB extensions. The live schema exposes preparation only when the execution environment supports it.

An empty send collects output. stdin answers a prompt or debugger; its text is queued exactly, so a line of input normally needs a trailing newline. Control interrupts or restarts the session. timeout_ms limits how long the call waits for an evaluation, rather than stopping the evaluation when that wait expires. Explicit preparation and control happen before that wait.

**Show:** This is a schematic field overview, not a call that supplies three languages at once. Keep the walkthrough brief; the next slides show control and input together. Code-free standalone preparation cannot carry nonempty stdin. Use stdin alone to answer an already active prompt.

**Sources:** [Send operation contract](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SEND_OPERATIONS.md) · [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::

## Interrupt or restart the session {#session-controls .deck-slide .literal-slide}
::: {.kicker}

Work with the session

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

Interrupt

:::

```python
console.send(control="interrupt")
```

::::

Request a cooperative stop. Keep the live workspace.

:::::

::::: {.right}

:::: {.code-panel}

::: {.panel-label}

Restart

:::

```python
console.send(control="restart")
```

::::

Replace the worker. Start with fresh in-memory state.

:::::

::::::

:::: {.code-panel}

::: {.panel-label}

R package development · start fresh, then load or test

:::

```python
console.send(control="restart", r="devtools::load_all()")
console.send(control="restart", r="devtools::test()")
```

::::

::: {.takeaway}

Restart keeps prepared dependencies and starts the supplied cell in the new session.

:::

::: {.notes}

**Say:** Interrupt requests SIGINT and preserves the live workspace. Code can catch or delay it, so this is cooperative. Restart retires the worker and starts a replacement. R objects, Python state, the managed DuckDB catalog, debugger state, and unread input are discarded; successfully prepared dependencies remain available.

A common package-development pattern is to restart and then call devtools::load_all() or devtools::test() in the same send. The package files remain on disk, while the operation runs in a fresh R session. Launch from the package directory for these examples. The two calls are alternatives.

**Show:** There are exactly two accepted control values. A control-only interrupt can overlap a pending send. If interrupt includes a new cell, that cell runs only after the previous evaluation has stopped; an uncooperative evaluation prevents it from running. Use the simpler restart example here to make ordering visible.

**Author / capture:** The restart-plus-test call was exercised against the local package fixture in the controls capture. The next slide captures restart plus a sourced script with queued input.

**Sources:** [Send operation contract](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SEND_OPERATIONS.md) · [Interruption and restart](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md#interruption)

:::

## Run an interactive script in a fresh session {#restart-input .deck-slide .literal-slide}
::: {.kicker}

Work with the session

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label .path-label}

examples/analyze.R

:::

```r
d <- read.csv("examples/measurements.csv")
group <- readline("Group to summarize: ")
values <- d$response[d$group == group]
print(summary(values))
```

::::

:::: {.code-panel}

::: {.panel-label}

One send call

:::

<!-- control-cell:restart-script -->
```python
console.send(
  control="restart",
  stdin="A\n",
  r='source("examples/analyze.R")'
)
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

One response · captured from Console

:::

```{r}
#| label: restart-input-output
#| eval: true
#| echo: false
#| comment: ""
console_protocol("controls/restart-script")
```

::::

:::::

::::::

::: {.takeaway}

Replace the worker → queue input → source the script.

:::

::: {.notes}

**Say:** An existing analysis script asks which group to summarize. The model can run it in a fresh session without editing out its interactive prompt. One request restarts the worker, queues A followed by a newline, and sources the script. When readline asks for the group, that input is already available. The server returns lifecycle notices and the script's output in one response.

The source path names a real file in examples/analyze.R. It reads the same measurements CSV used earlier in the talk. source does not automatically print each visible expression, so the script prints its summary explicitly. The newline is explicit because stdin does not add one.

**Show:** Read the script, then the call and response. Only the replacement receives same-call input. This is captured output, including the managed read notice; startup messages and prompt formatting can vary by runtime. The capture separately checks that an object from the old session is gone and the script selected group A. If the answer depends on output not yet seen, run the script first and answer its prompt with a later stdin-only call.

**Sources:** [Send operation order](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SEND_OPERATIONS.md#operations)

:::

## Default policy: read-only {#default-sandbox .deck-slide .policy-slide}
::: {.kicker}

Choose the worker’s permissions

:::

- Read host files.
- Write only in private temporary storage.
- Restrict direct network access.
- Apply the same policy to subprocesses.
- Retire owned processes and remove private storage on shutdown or restart.

::: {.takeaway}

Host reads are allowed. Deny sensitive paths explicitly.

:::

::: {.notes}

**Say:** The default is the read-only policy, available explicitly as :read-only. Before changing anything, here is what it provides: broad host reads, writes confined to private temporary storage, restricted direct networking, and the same policy for subprocesses. Project files and shared temporary directories are not generally writable.

The runner owns worker cleanup on restart and shutdown, removes its private temporary storage, and handles caller loss while it remains alive. On Linux, it waits for namespace retirement. On macOS, cleanup covers the owned process group and detached descendants the runner observes; it cannot promise to catch every orphaning race. Runner death has no independent recovery guarantee.

Read access and inherited environment variables can expose secrets, so configure explicit denials and environment controls when needed. Trusted package preparation is outside this worker policy. Session records, package caches, and intentionally written project files persist; temporary cleanup does not remove those.

**Show:** Explain these defaults before introducing a configuration file. Next, select a built-in policy at launch before adding project-specific permissions.

**Sources:** [Sandbox configuration](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md) · [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::

## Choose a built-in policy at launch {#builtin-policies .deck-slide .literal-slide}
::: {.kicker}

Choose the worker’s permissions

:::

:::: {.code-panel}

::: {.panel-label}

:read-only · read files, write only private scratch

:::

```bash
mcp-console serve -c extends=:read-only
```

::::

:::: {.code-panel}

::: {.panel-label}

:workspace · also allow writes in the project

:::

```bash
mcp-console serve -c extends=:workspace
```

::::

::: {.takeaway}

Both keep network restrictions and cleanup. Workspace mode protects project metadata.

:::

::: {.notes}

**Say:** The two named built-ins are :read-only and :workspace. The default gives read access and private scratch. Workspace mode additionally grants writes under the launch working directory while protecting .git, .agents, .codex, and .claude from writes. Networking and private-storage behavior remain restricted.

The -c option selects the policy for this launch. There are two built-in names in the current interface; external-sandbox and unrestricted are enforcement modes rather than additional named presets. The next slide puts the workspace choice in a project file and adds explicit read and deny entries.

**Show:** These are alternative launch commands. The leading colon is part of each built-in name. Project configuration is read first; -c overrides its extends field while retaining other configured fields. Demonstrate in a clean project when showing just the built-in baseline.

**Sources:** [Sandbox configuration](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md) · [Configuration layering](https://github.com/t-kalinowski/mcp-console/blob/main/docs/CONFIGURATION.md)

:::

## Extend a built-in policy for the project {#private-inputs .deck-slide .config-slide .code-dense}
::: {.kicker}

Choose the worker’s permissions

:::

:::::: {.two-col}

::::: {.left}

::: {.ownership-card}

./data: read-only.

./secrets and ./.env: denied.

Other workspace paths: writable, with protected project metadata.

:::

:::::

::::: {.right}

:::: {.code-panel}

::: {.panel-label .path-label}

.agents/console/config.yaml

:::

```yaml
extends: ":workspace"
sandbox:
  filesystem:
    entries:
      - path: {type: path, path: ./data}
        access: read
      - path: {type: path, path: ./secrets}
        access: deny
      - path: {type: path, path: ./.env}
        access: deny
```

::::

:::::

::::::
::: {.takeaway}

Set project permissions in .agents/console/config.yaml.

:::

::: {.notes}

**Say:** Console reads .agents/console/config.yaml in the launch directory once at startup. This is trusted user configuration. The default requires no file. extends: ":workspace" makes the previous launch choice part of the project configuration. For project work we can allow workspace writes, make the data directory read-only, and deny the secrets directory and .env. These are user-selected permissions captured when Console starts. The model can work inside that choice. The example shows why the sandbox needs more than a single on/off switch.

**Show:** Walk through the visible YAML from top to bottom. This is a complete alternative configuration, not a patch to concatenate with the preceding example.

**Author / capture:** Use pre-existing fixture paths. Native specificity is not array order. Linux has documented limits for nested deny/read reopenings; this simple example does not establish arbitrary cross-platform ACL equivalence.

**Sources:** [sandbox](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::

## Connect to a corporate data warehouse {#network-proxy .deck-slide .config-slide .code-dense}
::: {.kicker}

Choose the worker’s permissions

:::

:::::: {.two-col}

::::: {.left}

::: {.ownership-card}

Route worker traffic through the managed proxy.

Allow the corporate warehouse host.

Use its HTTPS database API, or a SOCKS-aware database connection.

:::

:::::

::::: {.right}

:::: {.code-panel}

::: {.panel-label .path-label}

.agents/console/config.yaml

:::

```yaml
extends: ":workspace"
sandbox:
  network: restricted
  proxy:
    enabled: true
    enableSocks5: true
    enableSocks5Udp: false
    allowUpstreamProxy: false
    dangerouslyAllowAllUnixSockets: false
    allowLocalBinding: false
    mode: full
    domains:
      warehouse.example.com: allow
      blocked.example.com: deny
```

::::

:::::

::::::
::: {.takeaway}

The client still supplies its usual database credentials.

:::

::: {.notes}

**Say:** For example, a corporate warehouse can expose an HTTPS query endpoint at the allowed host. A database using another TCP protocol needs a SOCKS-aware client or an explicitly configured SOCKS tunnel; a domain allowlist does not rewrite arbitrary database drivers. Network reachability and database authentication remain separate. For a database on a private IP, the pinned proxy requires a literal allowed IP matching the destination (or the broader local-binding exception); a hostname allowlist alone does not grant private-address access. The placeholder host is illustrative and was not contacted.

 This complete proxy configuration shows both the allowlist and the surrounding controls. The runner owns proxy routing, host normalization, and enforcement. A proxy object must include the required scalar fields; Console does not materialize omitted defaults for it.

**Show:** Walk through the visible YAML from top to bottom. This is a complete alternative configuration, not a patch to concatenate with the preceding example.

**Author / capture:** Example domains are placeholders, not actual services. All seven required scalar proxy fields are present. mode: full does not erase the domain allowlist. Review PROTOCOL.md for mode, upstream proxy, UDP and local-binding semantics.

**Sources:** [sandbox](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::

## Develop a local Shiny app {#network-listener .deck-slide .config-slide .code-dense}
::: {.kicker}

Choose the worker’s permissions

:::

:::::: {.two-col}

::::: {.left}

::: {.ownership-card}

allowLocalBinding: true

Run Shiny on the loopback interface.

Open `http://127.0.0.1:8000` in your browser.

:::

:::::

::::: {.right}

:::: {.code-panel}

::: {.panel-label .path-label}

.agents/console/config.yaml

:::

```yaml
extends: ":workspace"
sandbox:
  network: restricted
  proxy:
    enabled: true
    enableSocks5: true
    enableSocks5Udp: false
    allowUpstreamProxy: false
    dangerouslyAllowAllUnixSockets: false
    allowLocalBinding: true
    mode: full
    domains:
      api.example.com: allow
```

::::

:::::

::::::
::: {.takeaway}

`shiny::runApp("app", host = "127.0.0.1", port = 8000)`

:::

::: {.notes}

**Say:** This configuration supports local Shiny development: the worker can bind the loopback port while outbound requests use the managed proxy. Start an app at app/ with shiny::runApp and open its URL locally. A long-running app uses the wait, poll, and interrupt contract introduced earlier.

 An interactive local server can need a binding exception. The policy exposes that choice separately from the domain allowlist. Do not conflate permission to bind with permission to publish a remote service, create a tunnel, or expose every Unix socket.

**Show:** Walk through the visible YAML from top to bottom. This is a complete alternative configuration, not a patch to concatenate with the preceding example.

**Author / capture:** Validate the exact native behavior with the application used for the demo. This slide does not promise a Shiny viewer or automatic port forwarding. Keep all required proxy fields when changing a single flag.

**Sources:** [sandbox](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::

## Configure where the session runs {#execution-choice .deck-slide .story-slide}
::: {.kicker}

Choose the execution host

:::

:::: {.three-col}

::: {.story-card}

**Local host**

Beside your project, using this machine’s files and compute.

:::

::: {.story-card}

**SSH host**

On an existing remote machine, beside its data and compute.

:::

::: {.story-card}

**Docker container**

Inside a prepared image, with the mounts you select.

:::

::::

::: {.takeaway}

Execution targets plug into the same session interface.

:::

::: {.notes}

**Say:** The architecture separates session control from its execution target, so target adapters can support different environments. These cards show local, SSH, and Docker examples supported today. The session worker does not have to run beside the client. It can run locally, on an SSH host, or in Docker. The user selects the target at session launch, and the model still chooses R, Python, or SQL. Additional targets need an implementation of the relevant transport, environment, and lifecycle contracts; arbitrary environments do not work just by naming them.

Local and SSH targets support managed package preparation on the execution host. Docker uses the packages prepared in its image. The current implementation treats SSH and Docker as separate target modes.

**Sources:** [SSH execution](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SSH.md) · [Docker execution](https://github.com/t-kalinowski/mcp-console/blob/main/docs/DOCKER.md)

:::

## Send code to the remote host; receive results {#ssh-workflow .deck-slide .story-slide}
::: {.kicker}

SSH · Keep computation beside the existing data

:::

:::: {.two-col}

::: {.story-card}

**Your machine**

MCP client and Console server.

Send cells and controls.<br>Receive text and plots.<br>Keep session records.

:::

::: {.story-card}

**Remote host**

Sandboxed session worker.

Use the files already there.<br>Run R, Python, and SQL.<br>Prepare packages there.

:::

::::
::: {.takeaway}

The source data stays on the remote host. Results return over SSH.

:::

::: {.notes}

**Say:** This is an SSH session on an existing host. The project and source data are already there. Console sends code and control messages to that worker and receives its output, plots, and artifacts. It does not copy the project into a Docker sandbox.

Results can of course contain data selected by the submitted code. The point is that the dataset does not need to be staged locally to run the analysis. The remote host needs a compatible Console installation, runtime prerequisites, and the workspace. Both the account permissions and the worker sandbox apply.

**Sources:** [SSH execution](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SSH.md)

:::

## Point Console at the remote workspace {#ssh .deck-slide .config-slide .code-dense}
::: {.kicker}

SSH · Use an existing analysis host

:::

:::::: {.two-col}

::::: {.left}

::: {.ownership-card}

Remote workspace and relative paths.

Remote runner and resolver.

Local MCP server and recordings.

:::

:::::

::::: {.right}

:::: {.code-panel}

::: {.panel-label .path-label}

.agents/console/config.yaml

:::

```yaml
extends: ":workspace"
target:
  transport:
    kind: ssh
    host: analysis-host
  workspace: /srv/projects/analysis
sandbox:
  filesystem:
    entries:
      - path: {type: path, path: ./data}
        access: read
```

::::

:::::

::::::


::: {.notes}

**Say:** This is the local project configuration. analysis-host is an existing SSH destination, and /srv/projects/analysis is an existing directory on that host. The workspace baseline and the read-only data rule are materialized there. Then the client starts mcp-console serve in the usual way. Relative worker paths refer to the remote workspace; records stay with the controller.

**Show:** Walk through the visible YAML from top to bottom. This is a complete alternative configuration, not a patch to concatenate with the preceding example.

**Author / capture:** analysis-host must be a configured OpenSSH destination; /srv/projects/analysis and data must already exist. SSH+Docker composition is not supported in this revision.

**Sources:** [ssh](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SSH.md) · [sandbox](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::

## Run the session in a Docker container {#docker .deck-slide .config-slide .code-dense}
::: {.kicker}

Docker · One image and explicit mounts

:::

:::::: {.two-col}

::::: {.left}

::: {.ownership-card}

Image supplies runtimes and packages.

Project bind is explicitly writable.

Outer Docker boundary owns enforcement.

:::

:::::

::::: {.right}

:::: {.code-panel}

::: {.panel-label .path-label}

.agents/console/config.yaml

:::

```yaml
target:
  workspace: /workspace
  compute:
    kind: docker
    image: my-console:analysis
    pull: if_missing
    mounts:
      - source: .
        target: /workspace
        access: read_write
sandbox:
  filesystem: {kind: external-sandbox}
  network: enabled
```

::::

:::::

::::::


::: {.notes}

**Say:** This example selects a prepared image and binds the project into /workspace. external-sandbox delegates enforcement to the existing Docker boundary; the enabled network setting permits ordinary container networking. The writable bind is deliberate and exposes those host files to changes. Use the permissions the task needs. The next slide returns to what the model sees.

**Show:** Walk through the visible YAML from top to bottom. This is a complete alternative configuration, not a patch to concatenate with the preceding example.

**Author / capture:** Build my-console:analysis first. Source paths are interpreted by the selected Docker daemon. Native sandboxing inside Docker is a separate supported selection where host capabilities permit it; no privileged-mode fallback is implied.

**Sources:** [docker](https://github.com/t-kalinowski/mcp-console/blob/main/docs/DOCKER.md) · [sandbox](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::

## The agent uses the same interface on every host {#target-invariance .deck-slide}
::: {.kicker}

Local · SSH · Docker

:::

::: {.code-panel}

::: {.panel-label}

send

:::

```python
console.send(r="coef(fit)")
```

:::

::: {.target-line}

Local host &nbsp; / &nbsp; SSH &nbsp; / &nbsp; Docker

:::



::: {.notes}

**Say:** The model still sends a cell, receives results, and uses the same wait and control contract. It can use R, Python, and SQL whether the worker is beside the local project, on an SSH host, or in a container. The client above and the target below can change while the agent keeps the same send, result, and control interface. Available preparation capabilities differ by target and are reflected by the live tool schema.

**Show:** Return to one familiar call with three destinations beneath it. This is the bridge into client integrations: both the client above and execution target below can vary without changing the central idea.

**Sources:** [Implemented architecture](https://github.com/t-kalinowski/mcp-console/blob/main/docs/ARCHITECTURE.md) · [Canonical tool schema snapshot](https://github.com/t-kalinowski/mcp-console/blob/main/tests/snapshots/client_server/server/test_tools/initializes_and_lists_tools.yaml) · [Python clients and integrations](https://github.com/t-kalinowski/mcp-console/blob/main/docs/PYTHON.md)

:::

## Keep session logs and readable transcripts {#recording .deck-slide .literal-slide}
::: {.kicker}

Keep the work useful after the session

:::

:::: {.code-panel}

::: {.panel-label}

Files under one session directory

:::

```text
session/
├── internal/
│   └── events.jsonl     # Calls, results, request metadata
├── outputs/
│   └── call-000003.log  # Raw cell output
├── artifacts/
│   └── *.png           # Captured plots
├── transcript.md       # Readable session history
└── transcript.qmd      # Executable code cells
```

::::
::: {.takeaway}

Full logs and a readable presentation of the session.

:::

::: {.notes}

**Say:** Exploration should leave something useful behind. These files retain calls and results, raw output, plots, and two views of the session. Per-cell logs retain output up to the documented 1 GiB limit. The Markdown is a readable history. The Quarto source is a starting point for a report.

The structured event journal preserves request metadata supplied by the harness. It is not the entire MCP transport exchange; our capture script records that separately in wire.jsonl. Session records remain on the controller when the worker is remote.

**Show:** Walk down the tree. The comments are explanatory annotations, and the filenames reflect the captured layout.

**Sources:** [Architecture](https://github.com/t-kalinowski/mcp-console/blob/main/docs/ARCHITECTURE.md)

:::

## Turn the session into an editable starting point {#quarto-transcript .deck-slide .literal-slide .code-dense}
::: {.kicker}

Keep the work useful after the session

:::

:::: {.code-panel}

::: {.panel-label .path-label}

report.qmd · edited copy of the session transcript

:::

```{r}
#| label: quarto-document
#| eval: true
#| echo: false
#| comment: ""
#| results: asis
header <- console_capture_text("excerpts/quarto-header")
header <- sub("# Generated by MCP Console. Do not edit.\n", "", header, fixed = TRUE)
header <- sub("ir:\n", 'ir:\n  exclude-newer: "2026-09-17"\n', header, fixed = TRUE)
header <- sub("root.dir: [^\n]+", "root.dir: <project>", header)
header <- sub("(?s)  packages:.*\n---",
              "  packages: [tidyverse, reticulate] # list shortened\n---",
              header, perl = TRUE)
cat("````markdown\n", header, "\n",
    console_capture_text("excerpts/quarto-cell"), "\n````\n", sep = "")
```

::::

::: {.takeaway}

`ir render report.qmd` uses the recorded package snapshot date.

:::

::: {.notes}

**Say:** Add a snapshot date to the editable copy and render it with ir render report.qmd. ir consumes ir.exclude-newer to select dated default CRAN and Bioconductor repositories; declared Python requirements inherit the cutoff unless separately configured. This keeps subsequent resolution tied to the analysis date. It does not by itself pin every system dependency or arbitrary external source.

This is one Quarto document: front matter followed by an executable R chunk. Copy it into a report, edit the analysis, and render it in a suitable environment. Console owns the generated original; edit a copy.

Rendering runs the cells again. It does not replay interactive input, interrupts, or prior outputs. The data and any selected SQL connection still need to exist, and trusted rendering happens outside the worker sandbox. Remote projections default to eval:false.

**Show:** Read it from top to bottom as one source file. The snapshot date is an author-added setting, not a field emitted by the captured Console build. The root directory is shortened to <project>, the generated-file warning is omitted, and the dependency lists are shortened for display. The full source remains in captures/session-records/transcript.qmd. Dependencies reflect this captured build; they are not a complete record of inferred packages.

**Sources:** [ir Quarto integration](https://r-lib.github.io/ir/quarto.html) · [Architecture](https://github.com/t-kalinowski/mcp-console/blob/main/docs/ARCHITECTURE.md)

:::

## Use Console through your existing interface {#interfaces .deck-slide .story-slide}
::: {.kicker}

One execution system, several ways to connect

:::

:::: {.three-col}

::: {.story-card}

**MCP clients**

Compatible MCP stdio clients, including Codex and Claude Code.

:::

::: {.story-card}

**R**

ellmer through mcp.console.

:::

::: {.story-card}

**Python**

Sync and async clients.<br>chatlas.<br>OpenAI Responses and Agents SDK.<br>Anthropic.<br>Codex thread SDK.

:::

::::

::: {.notes}

**Say:** Console is an MCP server, so a compatible MCP client can use it directly. There is an ellmer tool in the R package, and Python offers direct clients and adapters for these frameworks. The next slides show a few registration examples.

The adapters preserve the live tool schema. Image support and client deadlines vary by interface, so check the chosen path before a demo. These interfaces are documented in the current source checkout; the installed PyPI 0.0.3 wheel used for runtime captures does not include the newer Python clients and extras.

**Sources:** [Python integrations](https://github.com/t-kalinowski/mcp-console/blob/main/docs/PYTHON.md) · [R interface](https://github.com/t-kalinowski/mcp-console/blob/main/r/README.md)

:::

## ellmer {#ellmer .deck-slide .integration}
::: {.kicker}

Use the client you already work in

:::

::: {.code-panel}

::: {.panel-label}

R · register one tool

:::

```r
library(ellmer)
library(mcp.console)

chat <- chat_openai()
chat$register_tool(console_tool())
chat$chat("Analyze measurements.csv with Console.")
```

:::

::: {.takeaway}

Your chat and provider. A separate persistent Console worker.

:::

::: {.notes}

**Say:** Give ellmer one slide. The R wrapper starts Console, reads its live tool schema, and presents it as an ellmer tool. It does not execute agent code in the R process running the chat. Provider choice remains with ellmer and the application. Keep the tool alive across the conversation so its session persists.

**Show:** Show only the registration example, with the registration line emphasized. No installation explanation here; that is the next section. The author notes include the asynchronous sequential-tool caveat from the R README.

**Sources:** [R package interface](https://github.com/t-kalinowski/mcp-console/blob/main/r/README.md) · [R tool wrapper implementation](https://github.com/t-kalinowski/mcp-console/blob/main/r/R/console-tool.R)

:::

## chatlas {#chatlas .deck-slide .integration}
::: {.kicker}

Use the client you already work in

:::

::: {.code-panel}

::: {.panel-label}

Python · preserve the server-provided schema

:::

```{.python code-line-numbers="2,5-7"}
from chatlas import ChatOpenAI
import mcp_console

chat = ChatOpenAI()
with mcp_console.MCPConsole() as console:
    tool = mcp_console.chatlas.tool(console)
    chat.set_tools([*chat.get_tools(), tool])
    chat.chat("Analyze measurements.csv with Console.")
```

:::

::: {.takeaway}

The same registration pattern in Python.

:::

::: {.notes}

**Say:** The chatlas adapter adds the same tool to an existing chat. Use the adapter and set_tools so the explicit server schema is preserved rather than inferred from a generic Python function annotation. The application still owns the chat loop and the Console connection lifetime.

**Show:** Keep the visual design identical to the ellmer slide. For a visual demo, verify image preservation on the chosen adapter path; the native MCP registration variant is included in the examples file.

**Sources:** [Python clients and integrations](https://github.com/t-kalinowski/mcp-console/blob/main/docs/PYTHON.md)

:::

## Register Console with your CLI client {#codex .deck-slide .integration}
::: {.kicker}

Use the client you already work in

:::

:::: {.code-panel}

::: {.panel-label}

Codex CLI

:::

```bash
codex mcp add console -- uvx mcp-console serve
```

::::
:::: {.code-panel}

::: {.panel-label}

Claude Code CLI · the same Console installation

:::

```bash
claude mcp add console -- uvx mcp-console serve
```

::::

::: {.takeaway}

Let `uvx` manage installation and pick up updates.

:::

::: {.notes}

**Say:** Both clients can launch the same Console server over MCP stdio. Run the registration command for the client you use. Client-side deadlines remain separate from Console’s wait timeout. In the ordinary uvx-managed path, uv resolves requirements using its registry cache and picks up newer versions as that cache expires. It does not promise a new registry lookup on every launch. An explicitly installed compatible tool can be reused instead, and version pins remain intentional. Keep those details in the notes; the visible point is that the launcher manages installation and updates.

**Show:** The launcher and server arguments are identical. The commands were checked against the installed CLI help; no configuration was changed.

**Sources:** [uv tools](https://docs.astral.sh/uv/concepts/tools/) · [Codex MCP](https://developers.openai.com/codex/mcp/) · [Claude Code MCP](https://code.claude.com/docs/en/mcp)

:::

## Now, installation from R {#install-r .deck-slide}
::: {.kicker}

Installation

:::

::: {.code-panel}

::: {.panel-label}

R package entry point

:::

```r
install.packages("mcp.console")
```

:::

::: {.takeaway}

Install the R interface. The application is obtained automatically when needed.

:::

::: {.notes}

**Say:** Only after the value and client examples are clear, show the R installation story. For presentation day, assume mcp.console is published to the intended R package repository. Installing the R package gives the interface; creating the tool resolves the Console executable when necessary. Do not say the PyPI download occurs during install.packages itself.

**Show:** Show one large line. This is explicitly an intended publication pathway: the currently inspected R README uses a GitHub installation command. Keep the present-day command in the author checklist, not as competing text on this slide.

**Author check before presenting:** Target-day assumption: mcp.console is published to an R repository usable by install.packages(). Current README documents pak::pak("github::t-kalinowski/mcp-console/r").

**Sources:** [R package interface](https://github.com/t-kalinowski/mcp-console/blob/main/r/README.md)

:::

## Python or a standalone MCP launch {#install-python .deck-slide .integration}
::: {.kicker}

Installation

:::

::: {.code-panel}

::: {.panel-label}

Python client

:::

```bash
pip install "mcp-console[client]"
```

:::

::: {.code-panel}

::: {.panel-label}

One-shot tool launch

:::

```bash
uvx mcp-console serve
```

:::

::: {.code-panel}

::: {.panel-label}

Persistent command installation

:::

```bash
uv tool install mcp-console
```

:::

::: {.notes}

**Say:** Python users choose the base package or the extra for their integration. MCP clients can launch through uvx, or use a persistent uv tool installation. These are alternative routes into the same distribution, not three installation steps. Framework extras avoid installing every SDK when only the executable is needed.

**Show:** Show three short alternatives with explicit labels. Mention prerequisites only as needed: Python packaging and supported native wheels are not the same thing as the language runtime installed on a selected execution host.

**Sources:** [Project README](https://github.com/t-kalinowski/mcp-console/blob/main/README.md) · [Python clients and integrations](https://github.com/t-kalinowski/mcp-console/blob/main/docs/PYTHON.md)

:::

## R, Python, and SQL, available for the work {#closing .deck-slide .hero .story-slide}
::: {.kicker}

Why Console exists

:::

::: {.big-statement}

Let the model choose the language.<br>You choose the execution host.<br>Give the agent an interactive workbench.

:::

::: {.takeaway}

Shared execution, managed dependencies, explicit permissions, and a record of the analysis.

:::

::: {.notes}

**Say:** The reason for Console is to give the model access to R, Python, and SQL together, with a common execution system underneath. That keeps R’s capabilities in reach, reduces duplicated runtime work, and lets us address the difficult parts in one place.

The examples—waiting without losing the run, compacting progress, and retaining output beyond the context window—show the level of detail we want across that system. The sandbox and execution target are part of the same design.

**Show:** End the main talk here. After the appendix divider, the following eight slides are optional development-practice material for questions.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [Sandbox configuration](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::

## Appendix {#appendix .deck-slide .hero .story-slide}
::: {.kicker}

Backup material

:::

::: {.big-statement}

How we test the execution system.

:::

::: {.takeaway}

Integration tests, snapshots, process boundaries, and installed products.

:::

::: {.notes}

**Say:** The main talk ends here. These slides are available for questions about the implementation and how its behavior is checked.

**Show:** Pause at this divider before entering the development material.

:::

## Rust implementation. Python integration tests. {#tests-python .deck-slide}
::: {.kicker}

Appendix · Development practices

:::

::::: {.two-col}

:::: {.left}

::: {.big-statement}

Implement in Rust

:::

::::

:::: {.right}

::: {.big-statement}

Exercise from Python

:::

::::

:::::



::: {.notes}

**Say:** The main integration suite is Python even though the application is Rust. It drives the built executable and observes the real process boundary. Small unit tests can still own pure parsing or validation policy; the claim is not that absolutely no Rust tests exist.

**Show:** Use two simple implementation/test boxes. Lead with why this matters: the test language can stay independent of implementation refactors while validating the product contract users and clients actually see.

**Sources:** [Boundary test guide](https://github.com/t-kalinowski/mcp-console/blob/main/tests/boundaries/README.md) · [Project README](https://github.com/t-kalinowski/mcp-console/blob/main/README.md)

:::

## Test the boundaries you intend to preserve {#test-boundaries .deck-slide}
::: {.kicker}

Appendix · Development practices

:::

<!-- inline-svg:assets/boundaries.svg -->
```{=html}
<div class="embedded-diagram">
<svg role="img" aria-label="Tests observe client-server, server-relay, relay-worker, and CLI boundaries." width="855pt" height="153pt"
 viewBox="0.00 0.00 854.80 152.80" xmlns="http://www.w3.org/2000/svg" xmlns:xlink="http://www.w3.org/1999/xlink">
<g id="diagram-11-graph0" class="graph" transform="scale(1 1) rotate(0) translate(14.4 138.4)">
<title>G</title>
<!-- client -->
<g id="diagram-11-node1" class="node">
<title>client</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M125,-124C125,-124 37,-124 37,-124 31,-124 25,-118 25,-112 25,-112 25,-90 25,-90 25,-84 31,-78 37,-78 37,-78 125,-78 125,-78 131,-78 137,-84 137,-90 137,-90 137,-112 137,-112 137,-118 131,-124 125,-124"/>
<text text-anchor="middle" x="81" y="-96.6" font-family="Arial" font-size="18.00" fill="#142d39">Test client</text>
</g>
<!-- server -->
<g id="diagram-11-node2" class="node">
<title>server</title>
<path fill="#d9eee9" stroke="#167d76" stroke-width="1.5" d="M397,-85C397,-85 299,-85 299,-85 293,-85 287,-79 287,-73 287,-73 287,-51 287,-51 287,-45 293,-39 299,-39 299,-39 397,-39 397,-39 403,-39 409,-45 409,-51 409,-51 409,-73 409,-73 409,-79 403,-85 397,-85"/>
<text text-anchor="middle" x="348" y="-57.6" font-family="Arial" font-size="18.00" fill="#142d39">Built server</text>
</g>
<!-- client&#45;&gt;server -->
<g id="diagram-11-edge1" class="edge">
<title>client&#45;&gt;server</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M137.23,-92.87C178.53,-86.79 235.22,-78.45 279.25,-71.97"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="279.73,-74.55 286.76,-70.86 278.96,-69.36 279.73,-74.55"/>
<text text-anchor="middle" x="224.5" y="-87.8" font-family="Arial" font-size="14.00" fill="#536b72">client_server</text>
</g>
<!-- relay -->
<g id="diagram-11-node3" class="node">
<title>relay</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M598,-85C598,-85 544,-85 544,-85 538,-85 532,-79 532,-73 532,-73 532,-51 532,-51 532,-45 538,-39 544,-39 544,-39 598,-39 598,-39 604,-39 610,-45 610,-51 610,-51 610,-73 610,-73 610,-79 604,-85 598,-85"/>
<text text-anchor="middle" x="571" y="-57.6" font-family="Arial" font-size="18.00" fill="#142d39">Relay</text>
</g>
<!-- server&#45;&gt;relay -->
<g id="diagram-11-edge2" class="edge">
<title>server&#45;&gt;relay</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M409.1,-62C445.31,-62 490.71,-62 524.04,-62"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="524.31,-64.63 531.81,-62 524.31,-59.38 524.31,-64.63"/>
<text text-anchor="middle" x="470.5" y="-65.8" font-family="Arial" font-size="14.00" fill="#536b72">server_relay</text>
</g>
<!-- worker -->
<g id="diagram-11-node4" class="node">
<title>worker</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M814,-85C814,-85 748,-85 748,-85 742,-85 736,-79 736,-73 736,-73 736,-51 736,-51 736,-45 742,-39 748,-39 748,-39 814,-39 814,-39 820,-39 826,-45 826,-51 826,-51 826,-73 826,-73 826,-79 820,-85 814,-85"/>
<text text-anchor="middle" x="781" y="-57.6" font-family="Arial" font-size="18.00" fill="#142d39">Worker</text>
</g>
<!-- relay&#45;&gt;worker -->
<g id="diagram-11-edge3" class="edge">
<title>relay&#45;&gt;worker</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M610.15,-62C643.33,-62 691.69,-62 728.33,-62"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="728.46,-64.63 735.96,-62 728.46,-59.38 728.46,-64.63"/>
<text text-anchor="middle" x="673" y="-65.8" font-family="Arial" font-size="14.00" fill="#536b72">relay_worker</text>
</g>
<!-- cli -->
<g id="diagram-11-node5" class="node">
<title>cli</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M150,-46C150,-46 12,-46 12,-46 6,-46 0,-40 0,-34 0,-34 0,-12 0,-12 0,-6 6,0 12,0 12,0 150,0 150,0 156,0 162,-6 162,-12 162,-12 162,-34 162,-34 162,-40 156,-46 150,-46"/>
<text text-anchor="middle" x="81" y="-18.6" font-family="Arial" font-size="18.00" fill="#142d39">CLI test process</text>
</g>
<!-- cli&#45;&gt;server -->
<g id="diagram-11-edge4" class="edge">
<title>cli&#45;&gt;server</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M162.23,-34.81C199.63,-40.31 243.65,-46.79 279.35,-52.04"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="279.03,-54.65 286.83,-53.14 279.79,-49.46 279.03,-54.65"/>
<text text-anchor="middle" x="224.5" y="-52.8" font-family="Arial" font-size="14.00" fill="#536b72">cli</text>
</g>
</g>
</svg>
</div>
```
<!-- /inline-svg -->

::: {.takeaway}

One primary owner per behavior; lower layers test their own contracts.

:::

::: {.notes}

**Say:** The tests mirror architectural boundaries: client_server, server_relay, relay_worker, and cli. Public behavior belongs at the outermost boundary that can usefully observe it. Private-boundary cases cover their own protocol seams rather than replicating every public message at every layer.

**Show:** Reuse the process topology with test-boundary names on the arrows. This ties the development method to the earlier architecture rather than presenting a directory tree with no explanation.

**Sources:** [Boundary test guide](https://github.com/t-kalinowski/mcp-console/blob/main/tests/boundaries/README.md)

:::

## YAML is what the reviewer reads {#yaml-transcripts .deck-slide .snapshot-slide}
::: {.kicker}

Appendix · Development practices

:::

::::: {.two-col}

:::: {.left}

::: {.code-panel}

::: {.panel-label}

Checked-in YAML · request

:::

```yaml
send:
  r: |
    cores <- parallel::detectCores()
    stopifnot(
      length(cores) == 1L,
      !is.na(cores),
      cores >= 1L
    )
    writeLines("R core detection available")
```

:::

::::

:::: {.right}

::: {.result-panel}

::: {.panel-label}

Checked-in YAML · result

:::

```yaml
result:
  content:
    - type: text
      text: |
        R core detection available
  isError: false
```

:::

::::

:::::

::: {.takeaway}

Readable code, arguments, output, and errors. Not escaped JSON strings.

:::

::: {.small-note}

Source excerpt: `detects_cpu_cores.yaml` · initialization document omitted.

:::

::: {.notes}

**Say:** The wire protocols are JSON-based, but the reviewable test transcripts are YAML. This is an actual checked-in snapshot excerpt: the test executes R, asserts that CPU detection returns a valid result, and prints a stable message. Humans can review the code and result without reading escaped JSON strings or snapshotting a machine-specific core count. YAML is the human review surface, not the production transport.

**Show:** Display the request and result from `tests/snapshots/client_server/r/test_runtime/detects_cpu_cores.yaml` side by side. The initial shared-handshake document is omitted, but the displayed request and result content come from the fetched source. This fixture was not rerun in the slide-build environment; do not present it as a new test execution.

**Sources:** [Boundary test guide](https://github.com/t-kalinowski/mcp-console/blob/main/tests/boundaries/README.md) · [Checked-in CPU detection snapshot](https://github.com/t-kalinowski/mcp-console/blob/main/tests/snapshots/client_server/r/test_runtime/detects_cpu_cores.yaml)

:::

## Normalize noise, not behavior {#test-normalization .deck-slide}
::: {.kicker}

Appendix · Development practices

:::

::::: {.two-col}

:::: {.left}

::: {.code-panel}

::: {.panel-label}

Unstable details

:::

```text
/tmp/run-8fd31/work.csv
worker pid: 27186
[waiting for stdin]
```

:::

::::

:::: {.right}

::: {.code-panel}

::: {.panel-label}

Schematic normalization

:::

```text
<WORKSPACE>/work.csv
worker pid: <PID>
[waiting for stdin]
```

:::

::::

:::::

::: {.takeaway}

Keep meaningful wording, ordering, and failure distinctions visible.

:::

::: {.notes}

**Say:** Temporary paths, process identities, and similar unstable details should not obscure behavioral review. Normalize explicitly and narrowly. Preserve the fields, output, ordering, and failure distinctions the contract is meant to protect. The actual project normalizers and snapshot metadata are more specific than this schematic example.

**Show:** Use before/after panels with only incidental details changed. Call out that normalization is not a license to delete inconvenient evidence from a test.

**Sources:** [Boundary test guide](https://github.com/t-kalinowski/mcp-console/blob/main/tests/boundaries/README.md)

:::

## Synchronize; do not sleep and hope {#test-determinism .deck-slide}
::: {.kicker}

Appendix · Development practices

:::

::: {.big-statement}

Start → checkpoint → release → assert

:::

::: {.takeaway}

Explicit gates for output, cancellation, restart, and cleanup.

:::

::: {.notes}

**Say:** Concurrency and liveness tests need causal synchronization. Fixtures use gates and checkpoints so the test knows when the relevant state has actually been reached. Arbitrary sleeps are not proof that output was drained, an interrupt was delivered, or an owned resource was retired.

**Show:** Show one deterministic sequence and connect it to the timing and shutdown contracts from the talk. This is a development-practice slide, not an implementation recipe for every fixture.

**Sources:** [Boundary test guide](https://github.com/t-kalinowski/mcp-console/blob/main/tests/boundaries/README.md)

:::

## Portable behavior; real capability checks {#test-capabilities .deck-slide}
::: {.kicker}

Appendix · Development practices

:::

::::: {.two-col}

:::: {.left}

::: {.big-statement}

Shared cases<br>direct + sandboxed

:::

::::

:::: {.right}

::: {.big-statement}

Capability fixtures<br>SSH · Docker · SBX

:::

::::

:::::

::: {.takeaway}

Fake peers establish protocol contracts; real targets establish target behavior.

:::

::: {.notes}

**Say:** Reuse portable cases across execution modes and centralize capability discovery. A skipped target fixture is not target validation. Deterministic peers can establish orchestration behavior, but they cannot establish real container or microVM cleanup. Keep real-target evidence distinct from simulated protocol coverage.

**Show:** Use the two testing levels rather than an all-green platform matrix. Include lifecycle and security assertions where a textual snapshot alone cannot represent the guarantee.

**Sources:** [Boundary test guide](https://github.com/t-kalinowski/mcp-console/blob/main/tests/boundaries/README.md)

:::

## Snapshots are generated evidence {#snapshot-review .deck-slide}
::: {.kicker}

Appendix · Development practices

:::

::: {.code-panel}

::: {.panel-label}

Run, regenerate, review

:::

```bash
scripts/test --update BOUNDARY/SUITE::CASE

git diff -- tests/snapshots/
```

:::



::: {.notes}

**Say:** Generate transcripts from running tests, then review the resulting diff. Do not hand-edit the expectation to make a test pass. The value of readable YAML is that review can focus on the actual changed behavior, while assertions still enforce facts that a snapshot cannot show.

**Show:** Show the update command and a diff review, not a large wall of green tests. The selector is schematic; replace BOUNDARY/SUITE::CASE with a real selected fixture for a live development demonstration.

**Sources:** [Boundary test guide](https://github.com/t-kalinowski/mcp-console/blob/main/tests/boundaries/README.md)

:::

## Test the installed product {#installed-product-tests .deck-slide}
::: {.kicker}

Appendix · Development practices

:::

::::: {.two-col}

:::: {.left}

::: {.big-statement}

Source build<br>and wheels

:::

::::

:::: {.right}

::: {.big-statement}

R / Python clients<br>and native bundle

:::

::::

:::::

::: {.takeaway}

Packaging, launch, schemas, images, shutdown, and cleanup are part of the contract.

:::

::: {.notes}

**Say:** The delivered product includes the executable, companion binaries, Python interfaces, the R wrapper, and launch/lifetime behavior. Running a development binary alone does not establish that the installed bundle or each adapter works. Integration examples and packaging checks should be part of the release evidence, with actual target validation reported separately.

**Show:** Finish the addendum by connecting the install story to the test strategy. Avoid claiming every proposed integration test already exists: distinguish observed project coverage from the release checklist in the companion author notes.

**Sources:** [Boundary test guide](https://github.com/t-kalinowski/mcp-console/blob/main/tests/boundaries/README.md) · [Release and installed bundle](https://github.com/t-kalinowski/mcp-console/blob/main/RELEASE.md) · [Python clients and integrations](https://github.com/t-kalinowski/mcp-console/blob/main/docs/PYTHON.md) · [R package interface](https://github.com/t-kalinowski/mcp-console/blob/main/r/README.md)

:::
