---
# Canonical source. Notes use Quarto/reveal.js Speaker View (S).
# No automatic title slide: the first explicit ## heading is slide 1.
# Default: native R output/plots. Optional: captured MCP output via output_source:mcp.
title: ""
pagetitle: "MCP Console · interactive work, small interface"
engine: knitr
params:
  output_source: r
format:
  revealjs:
    slide-level: 2
    output-file: mcp-console.html
    show-notes: false
    width: 1600
    height: 900
    margin: 0
    center: false
    transition: none
    slide-number: c/t
    embed-resources: true
    code-copy: false
    code-line-numbers: false
    controls: true
    progress: true
    css: styles.css
execute:
  eval: false
  echo: false
  warning: false
  message: false
---

```{r}
#| label: render-setup
#| eval: true
#| include: false
source("examples/render-helpers.R", local = knitr::knit_global())
```

## MCP Console {#interactive-work .deck-slide .hero}
::: {.kicker}

The model-facing interface

:::

::: {.big-statement}

Interactive work.<br>A small interface.

:::

::: {.language-line}

R · Python · SQL

:::

::: {.takeaway}

Token-conscious tool design. Rich runtime capabilities.

:::

::: {.notes}

**Say:** This audience already knows how to give a model an evaluator. Skip the history of code execution tools. MCP Console is a place for models to work interactively: inspect a result, change direction, keep state, and continue. The subject of this talk is the model-facing interface, not a new chat UI.

**Show:** Open directly on the product name and the two-line thesis. No definition of MCP, shell-versus-notebook comparison, or installation instructions. Advance into the actual tool contract after one brief framing slide.

**Sources:** [Project README](https://github.com/t-kalinowski/mcp-console/blob/main/README.md) · [Canonical tool schema snapshot](https://github.com/t-kalinowski/mcp-console/blob/main/tests/snapshots/client_server/server/test_tools/initializes_and_lists_tools.yaml)

:::



## Small surface. Deep runtime. {#small-surface .deck-slide}
::: {.kicker}

The model-facing interface

:::

<!-- inline-svg:assets/surface.svg -->
```{=html}
<div class="embedded-diagram">
<svg role="img" aria-label="The LLM client calls Console to use runtime capabilities." width="705pt" height="97pt"
 viewBox="0.00 0.00 705.00 97.00" xmlns="http://www.w3.org/2000/svg" xmlns:xlink="http://www.w3.org/1999/xlink">
<g id="diagram-1-graph0" class="graph" transform="scale(1 1) rotate(0) translate(14.4 82.32)">
<title>G</title>
<!-- model -->
<g id="diagram-1-node1" class="node">
<title>model</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M100.68,-57.42C100.68,-57.42 12,-57.42 12,-57.42 6,-57.42 0,-51.42 0,-45.42 0,-45.42 0,-22.5 0,-22.5 0,-16.5 6,-10.5 12,-10.5 12,-10.5 100.68,-10.5 100.68,-10.5 106.68,-10.5 112.68,-16.5 112.68,-22.5 112.68,-22.5 112.68,-45.42 112.68,-45.42 112.68,-51.42 106.68,-57.42 100.68,-57.42"/>
<text xml:space="preserve" text-anchor="middle" x="56.34" y="-27.36" font-family="Arial" font-size="18.00" fill="#142d39">LLM client</text>
</g>
<!-- send -->
<g id="diagram-1-node2" class="node">
<title>send</title>
<path fill="#d9eee9" stroke="#167d76" stroke-width="1.5" d="M295.61,-57.42C295.61,-57.42 171.68,-57.42 171.68,-57.42 165.68,-57.42 159.68,-51.42 159.68,-45.42 159.68,-45.42 159.68,-22.5 159.68,-22.5 159.68,-16.5 165.68,-10.5 171.68,-10.5 171.68,-10.5 295.61,-10.5 295.61,-10.5 301.61,-10.5 307.61,-16.5 307.61,-22.5 307.61,-22.5 307.61,-45.42 307.61,-45.42 307.61,-51.42 301.61,-57.42 295.61,-57.42"/>
<text xml:space="preserve" text-anchor="middle" x="233.65" y="-27.36" font-family="Arial" font-size="18.00" fill="#142d39">console.send()</text>
</g>
<!-- model&#45;&gt;send -->
<g id="diagram-1-edge1" class="edge">
<title>model&#45;&gt;send</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M113.29,-33.96C124.64,-33.96 136.81,-33.96 148.88,-33.96"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="148.76,-36.59 156.26,-33.96 148.76,-31.34 148.76,-36.59"/>
</g>
<!-- runtime -->
<g id="diagram-1-node3" class="node">
<title>runtime</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M663.79,-67.92C663.79,-67.92 366.61,-67.92 366.61,-67.92 360.61,-67.92 354.61,-61.92 354.61,-55.92 354.61,-55.92 354.61,-12 354.61,-12 354.61,-6 360.61,0 366.61,0 366.61,0 663.79,0 663.79,0 669.79,0 675.79,-6 675.79,-12 675.79,-12 675.79,-55.92 675.79,-55.92 675.79,-61.92 669.79,-67.92 663.79,-67.92"/>
<text xml:space="preserve" text-anchor="middle" x="515.2" y="-37.86" font-family="Arial" font-size="18.00" fill="#142d39">Runtime capabilities</text>
<text xml:space="preserve" text-anchor="middle" x="515.2" y="-16.86" font-family="Arial" font-size="18.00" fill="#142d39">objects · packages · DB connections</text>
</g>
<!-- send&#45;&gt;runtime -->
<g id="diagram-1-edge2" class="edge">
<title>send&#45;&gt;runtime</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M308.09,-33.96C319.46,-33.96 331.58,-33.96 344.04,-33.96"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="343.87,-36.59 351.37,-33.96 343.87,-31.34 343.87,-36.59"/>
</g>
</g>
</svg>
</div>
```
<!-- /inline-svg -->

::::: {.two-col}

:::: {.left}

::: {.big-statement}

Low ceremony

:::

::::

:::: {.right}

::: {.big-statement}

Room for precision

:::

::::

:::::

::: {.takeaway}

Ordinary code works; stronger models can use more of the contract.

:::

::: {.notes}

**Say:** The goal is token efficiency without making the environment artificially limited. Routine calls stay small. Package loading, object interchange, and database selection live naturally in the runtime; requirements and lifecycle controls are available when the model needs to be deliberate. Do not claim a measured speedup or superiority across model families.

**Show:** Use the three-box diagram. The same tool reaches a much larger runtime surface. This is the organizing idea for the sequence: introduce a field only when a concrete interaction needs it.

**Sources:** [Canonical tool schema snapshot](https://github.com/t-kalinowski/mcp-console/blob/main/tests/snapshots/client_server/server/test_tools/initializes_and_lists_tools.yaml) · [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [Requirements and environments](https://github.com/t-kalinowski/mcp-console/blob/main/docs/REQUIREMENTS.md)

:::



## Start with a cell {#send-r .deck-slide .literal-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg new-arg">r</span><span class="sig-name">)</span></div>
:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

LLM client → Console

:::

<!-- cells:fit -->
```python
console.send(r="""
  d <- read.csv("examples/measurements.csv")
  fit <- lm(response ~ temperature + group,
            data = d)
  fit
""")
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

R output

:::

```{r}
#| label: fit-output
#| eval: true
#| echo: false
#| comment: ""
console_text("fit")
```

::::

:::::

::::::
::: {.takeaway}

The value returned by the cell is the fitted model.

:::

::: {.notes}

**Say:** The final expression is fit, so R prints its normal lm summary. The agent receives useful output immediately, not prose that stands in for output. This is a real, persistent R workspace; ordinary cells do not require another object-management API.

**Show:** Show the exact submitted R code beside the printed lm object. The code and renderer both read the same cell definition. The fit is reused on the next slide.

**Author / capture:** Default preview evaluates this cell in native R through knitr, not in MCP Console. With -P output_source:mcp, the panel reads the literal captured Console text. No simulated numerical output is committed.

**Sources:** [runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [send](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SEND_OPERATIONS.md)

:::


## The next call uses the same fit {#live-state .deck-slide .literal-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="sig-name">)</span></div>

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

Call 1 · coefficients

:::

<!-- cells:coefficients -->
```python
console.send(r='coef(fit)')
```

::::
:::: {.code-panel}

::: {.panel-label}

Call 2 · predictions

:::

<!-- cells:predict -->
```python
console.send(r='head(predict(fit, d), 3)')
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

Response 1

:::

```{r}
#| eval: true
#| echo: false
#| comment: ""
console_text("coefficients")
```

::::
:::: {.result-panel}

::: {.panel-label}

Response 2

:::

```{r}
#| eval: true
#| echo: false
#| comment: ""
console_text("predict")
```

::::

:::::

::::::

::: {.notes}

**Say:** Read the coefficients and then the first three predictions from the existing fit. Persistence is demonstrated by the relationship between these calls, not by commentary placed in an output block.

**Show:** Keep coefficients and predictions in the result panel and all interpretation beneath it. Native Quarto re-renders these values from the preceding model; capture mode displays actual tool results.

**Sources:** [runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::

## Plots are results, not another tool {#r-plots .deck-slide .literal-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="sig-name">)</span></div>
:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

LLM client → Console

:::

<!-- cells:r-plot -->
```python
console.send(r="""
  plot(d$temperature, d$response,
       xlab = "Temperature",
       ylab = "Response")
""")
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

R graphics output

:::

```{r}
#| label: r-plot-output
#| eval: true
#| echo: false
#| comment: ""
#| fig-width: 7
#| fig-height: 5.2
#| dpi: 144
#| dev: png
#| out-width: "100%"
console_plot("r-plot")
```

::::

:::::

::::::
::: {.takeaway}

The managed R graphics device returns an image.

:::

::: {.notes}

**Say:** This is the base-R plot produced by this exact plot() call. Plotting is part of the runtime contract, not another tool. All drawing for a managed plot belongs in the same cell.

**Show:** Let R render this figure during quarto preview. Do not substitute a Matplotlib graphic. Capture mode uses the PNG returned by Console instead.

**Author / capture:** Default uses the displayed base-R cell through knitr; plot capture mode reads captures/r-plot-01.png. Neither a stock image nor the previous illustrative SVG is used.

**Sources:** [runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::


## A wait timeout is not an execution deadline {#wait-timeout .deck-slide .literal-slide .code-dense}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg new-arg">timeout_ms</span><span class="sig-name">)</span></div>
:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

LLM client → Console · running cell

:::

<!-- cells:progress -->
```python
console.send(r=r"""
  for (i in 1:20) {
    cat(sprintf("\rfit [%s%s] %3d%%",
        strrep("=", i), strrep(" ", 20 - i), i * 5))
    flush.console()
    Sys.sleep(0.15)
  }
  cat("\n")
""", timeout_ms=450)
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

Console → text · captured response

:::

```{r}
#| label: progress-first
#| eval: true
#| echo: false
#| comment: ""
console_protocol("progress")
```

::::
::: {.small-note}

The percentage depends on when the wait expires.

:::

:::::

::::::

::: {.notes}

**Say:** A response can end while the cell remains active. Here the R code repeatedly rewrites one progress line, and timeout_ms returns control before the loop completes. This is an evaluation wait budget, not a deadline that kills the computation.

**Show:** Read both the compact progress line and the exact running notice. The panel reads the captured response; call boundaries remain timing-dependent.

**Author / capture:** Warm up the session before capturing this example. Do not promise exactly 15% at 450 ms. The local capture script retains actual responses rather than asserting fixed percentages.

**Sources:** [send](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SEND_OPERATIONS.md) · [runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::


## Polling advances the same computation {#poll .deck-slide .literal-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="sig-name">)</span></div>
:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

LLM client → Console · collect new output

:::

<!-- cells:poll-next,poll-final -->
```python
console.send(timeout_ms=600)

console.send(timeout_ms=5000)
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

Console → text · after 600 ms

:::

```{r}
#| label: poll-first
#| eval: true
#| echo: false
#| comment: ""
console_protocol("poll-next")
```

::::

:::: {.result-panel}

::: {.panel-label}

Console → text · after 5000 ms

:::

```{r}
#| label: poll-last
#| eval: true
#| echo: false
#| comment: ""
console_protocol("poll-final")
```

::::
:::::

::::::
::: {.takeaway}

One current progress line per collected interval; never rerun the loop.

:::

::: {.notes}

**Say:** These are separate responses to later polls, not a transcript dumped into one response. Each poll contains newly collected output. Carriage-return redraws within one response interval collapse to its current line; a later interval can return the next current line.

**Show:** Point to the two poll calls and their corresponding results. The last response has real output and no invented [done] appended. Emphasize that empty send() is also a poll.

**Author / capture:** Both rendering modes read the two literal captured responses. The wire exchange records the 600 ms poll followed by the 5000 ms poll. Exact progress fractions are not a public API guarantee.

**Sources:** [runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [send](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SEND_OPERATIONS.md)

:::


## Progress redraws do not become a wall of text {#progress-redraws .deck-slide .literal-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="sig-name">)</span></div>
:::: {.code-panel}

::: {.panel-label}

LLM client → Console

:::

<!-- cells:compact -->
```python
console.send(r='cat("download 0%\\rdownload 50%\\rdownload 100%\\n")')
```

::::
:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

Emitted characters · escaped representation

:::

```text
download 0%\rdownload 50%\rdownload 100%\n
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

Console → text

:::

```{r}
#| label: compacted-output
#| eval: true
#| echo: false
#| comment: ""
console_protocol("compact")
```

::::

:::::

::::::
::: {.takeaway}

The raw log keeps the emitted stream; the model gets the compacted view.

:::

::: {.notes}

**Say:** A carriage return replaces the current progress frame rather than adding another visible line. This deterministic example is separate from timing-sensitive polling: all three writes are in one short cell. Raw logs preserve the original bytes.

**Show:** Show the escaped raw characters on the left and the literal compacted result on the right. Both boxes identify their representation explicitly; the explanation stays outside them.

**Sources:** [runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::


## Interrupt without replacing the runtime {#interrupt .deck-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg new-arg">control</span><span class="sig-name">)</span></div>

::: {.code-panel}

::: {.panel-label}

send

:::

```python
console.send(control="interrupt")
```

:::

::: {.big-statement}

Stop the work.<br>Keep the workspace.

:::

::: {.small-note}

Cooperative interrupt; restart is the stronger operation.

:::

::: {.notes}

**Say:** Add control. Interrupt requests the runtime or active resolver to stop its current work without deliberately replacing the worker. It is cooperative: code can delay or catch interruption. Keeping this distinct from restart makes state loss explicit rather than an accidental side effect of a timeout.

**Show:** Highlight only control in the signature. The single-line call is the main visual, followed by the two-line state-preservation message. Do not imply every interrupt must succeed immediately.

**Sources:** [Send operation contract](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SEND_OPERATIONS.md) · [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::



## Restart gives you a fresh workspace {#restart .deck-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="sig-name">)</span></div>

:::: {.code-panel}

::: {.panel-label}

Restart the worker

:::

```python
console.send(control="restart")
```

::::
:::::: {.two-col}

::::: {.left}

::: {.ownership-card}

Starts fresh

R and Python objects

In-memory SQL state

Debugger and pending input

:::

:::::

::::: {.right}

::: {.ownership-card}

Stays available

Prepared requirements

Session recordings

:::

:::::

::::::

::: {.notes}

**Say:** Restart replaces the worker. Objects, the in-memory DuckDB catalog, debugger state, and unread input are cleared. The server retains successfully prepared requirements and recordings. A later call starts fresh.

**Sources:** [Send operations](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SEND_OPERATIONS.md)

:::

## Inspect state after an error {#errors .deck-slide .literal-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="sig-name">)</span></div>

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

Call 1

:::

<!-- cells:error -->
```python
console.send(r="""
  checkpoint <- 42
  stop("inspect me")
""")
```

::::
:::: {.code-panel}

::: {.panel-label}

Call 2

:::

<!-- cells:checkpoint -->
```python
console.send(r='checkpoint')
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

Response 1

:::

```{r}
#| eval: true
#| echo: false
#| comment: ""
console_protocol("error")
```

::::
:::: {.result-panel}

::: {.panel-label}

Response 2

:::

```{r}
#| eval: true
#| echo: false
#| comment: ""
console_protocol("checkpoint")
```

::::

:::::

::::::

::: {.notes}

**Say:** An ordinary R error or Python exception does not normally destroy the worker. Also, cells are not transactions: assignments and other effects before the error can remain. That is a useful, explicit contract for an agent deciding whether to inspect, repair, retry, or restart.

**Show:** Show a failed cell followed by a successful inspection of the earlier assignment. No new argument appears. Keep the two responses visibly distinct from one response block.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [Canonical tool schema snapshot](https://github.com/t-kalinowski/mcp-console/blob/main/tests/snapshots/client_server/server/test_tools/initializes_and_lists_tools.yaml)

:::

## Interactive input is a separate channel {#stdin .deck-slide .literal-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg new-arg">stdin</span><span class="sig-name">)</span></div>
:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

LLM client → Console · start a read

:::

<!-- cells:prompt,prompt-answer -->
```python
console.send(r="""
  name <- readline("name> ")
  name
""")

console.send(stdin='Ada\n')
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

Console → text · first response

:::

```{r}
#| label: prompt-output
#| eval: true
#| echo: false
#| comment: ""
console_protocol("prompt")
```

::::
:::: {.result-panel}

::: {.panel-label}

Console → text · input response

:::

```{r}
#| label: prompt-answer-output
#| eval: true
#| echo: false
#| comment: ""
console_protocol("prompt-answer")
```

::::

:::::

::::::
::: {.takeaway}

Code and exact input bytes are separate fields.

:::

::: {.notes}

**Say:** The running cell asks for a line. Console exposes the input request, and the next call supplies the newline explicitly. The following result is the value returned by the same suspended evaluation.

**Show:** Keep the two Console responses in separate result panels. Both rendering modes read the complete real responses to these exact calls.

**Sources:** [runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [send](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SEND_OPERATIONS.md)

:::


## The debugger uses the same field {#debugger .deck-slide .literal-slide .code-dense}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="sig-name">)</span></div>
:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

LLM client → Console · enter a live function

:::

<!-- cells:browser-start -->
```python
console.send(r="""
  inspect_mean <- function(x) {
    browser()
    mean(x)
  }

  inspect_mean(c(1, 2, 3))
""")
```

::::


:::: {.result-panel}

::: {.panel-label}

Console → text · debugger prompt

:::

```{r}
#| label: browser-start
#| eval: true
#| echo: false
#| comment: ""
console_protocol("browser-start")
```

::::

:::::

::::: {.right}

:::: {.code-panel}

::: {.panel-label}

Model → stdin · inspect, then continue

:::

<!-- cells:browser-x,browser-continue -->
```python
console.send(stdin='x\n')

console.send(stdin='c\n')
```

::::
:::: {.result-panel}

::: {.panel-label}

Console → text · inspect x

:::

```{r}
#| label: browser-x
#| eval: true
#| echo: false
#| comment: ""
console_protocol("browser-x")
```

::::

:::: {.result-panel}

::: {.panel-label}

Console → text · continue

:::

```{r}
#| label: browser-continue
#| eval: true
#| echo: false
#| comment: ""
console_protocol("browser-continue")
```

::::
:::::

::::::
::: {.takeaway}

browser() is visible in the function body. Input inspects that live frame.

:::

::: {.notes}

**Say:** Stop inside inspect_mean, inspect x in that function frame, then continue to the mean. The same stdin field used for readline handles the debugger. A prompt is not new top-level code, and the session is not idle while it is waiting for debugger input.

**Show:** Use the multiline function at the left; browser() gets its own line. The two input responses include the returned R values and any new debugger prompt or waiting notice.

**Author / capture:** The panels read the complete captures/browser-start.txt, captures/browser-x.txt, and captures/browser-continue.txt responses. Do not run browser() in the knitr rendering process.

**Sources:** [runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [send](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SEND_OPERATIONS.md)

:::


## Keep model context bounded {#bounded-output .deck-slide .literal-slide .code-dense}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="sig-name">)</span></div>

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

Call

:::

<!-- cells:flood -->
```python
console.send(r=r"""
  for (i in seq_len(10000)) {
    cat(sprintf("row %05d\n", i))
  }
  cat("fit complete\n")
""")
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

One tool response · shortened for display

:::

```text
row 00001
row 00002
row 00003

[output preview: middle omitted;
 raw output retained at
 .agents/console/sessions/<run>/
 outputs/call-000008.log]

row 09999
row 10000
fit complete
```

::::

:::::

::::::

::: {.notes}

**Say:** One call returns one response. The text keeps a beginning and a recent tail, with an omission notice and a path to retained output between them. The text budget is 8 KiB per result; images have a separate allowance.

**Show:** This is one stylized response. The middle notice and path are shortened, and most preview lines are omitted to fit the slide. The complete literal response is in captures/flood.txt.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::

## Capture output beyond language hooks {#retained-output .deck-slide .literal-slide .code-dense}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="sig-name">)</span></div>

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

R · output from a forked child

:::

<!-- cells:r-fork -->
```python
console.send(r=r"""
  emit <- inline::cfunction(
    body = 'write(1, "native output\\n", 14);
            return R_NilValue;',
    includes = "#include <unistd.h>"
  )
  job <- parallel::mcparallel(emit())
  invisible(parallel::mccollect(job))
""")
```

::::
:::: {.result-panel}

::: {.panel-label}

Console response

:::

```{r}
#| eval: true
#| echo: false
#| comment: ""
console_protocol("r-fork")
```

::::

:::::

::::: {.right}

:::: {.code-panel}

::: {.panel-label}

Python · write directly to stdout

:::

<!-- cells:python-fd -->
```python
console.send(python=r"""
import os
_ = os.write(1, b"hello directly on fd 1\n")
""")
```

::::
:::: {.result-panel}

::: {.panel-label}

Console response

:::

```{r}
#| eval: true
#| echo: false
#| comment: ""
console_protocol("python-fd")
```

::::

:::::

::::::

::: {.notes}

**Say:** inline compiles a small C function that writes directly to file descriptor 1. The R call invokes it in a forked child and waits for that child to finish. The inline namespace was prepared before the displayed call; its compiler setup is preserved in the capture. The Python call writes bytes directly to file descriptor 1, bypassing Python’s sys.stdout object. Console captures both through the session’s output streams. Both examples bypass the ordinary language output hooks. The R fork example was captured on macOS and uses the Unix-only parallel::mcparallel API.

**Show:** Each example has its own actual captured response. Longer text is retained in per-cell logs, up to the retention limit, with a bounded preview returned to the client. The examples finish their children before returning.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::

## Package loading is a runtime capability {#r-resolution .deck-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="sig-name">)</span></div>

::: {.code-panel}

::: {.panel-label}

send

:::

```python
console.send(r="""
  library(data.table)
  as.data.table(d)[, .(n = .N), by = group]
""")
```

:::

::: {.takeaway}

Missing package → prepare → resume the original operation.

:::



::: {.notes}

**Say:** Only now introduce package resolution, after wait timeouts and the interactive contract. The simple model path is ordinary library() or pkg::fun() use. When execution reaches a supported missing-package operation, Console prepares the package and resumes that operation in the live worker. Quoted or unreachable code does not trigger a source-scanning installer.

**Show:** Keep the signature unchanged and show ordinary R. This is the first example of pushing capability into the runtime rather than expanding the tool schema. Resolver work may outlast the tool wait, using the timeout and polling behavior already explained.

**Sources:** [Requirements and environments](https://github.com/t-kalinowski/mcp-console/blob/main/docs/REQUIREMENTS.md) · [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::



## Pin what matters; leave the rest unpinned {#requirements .deck-slide .literal-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="arg new-arg">requirements</span><span class="sig-name">)</span></div>
:::: {.code-panel}

::: {.panel-label}

LLM client → Console

:::

<!-- cells:requirements -->
```python
console.send(r="""
  library(data.table)
  library(dtplyr)
  packageVersion("data.table")
""",
    requirements={'r': ['data.table==1.17.8', 'dtplyr']})
```

::::
:::::: {.two-col}

::::: {.left}

::: {.ownership-card}

`data.table==1.17.8`

An exact version requirement.

:::

:::::

::::: {.right}

::: {.ownership-card}

`dtplyr`

No explicit version pin.

:::

:::::

::::::

::: {.notes}

**Say:** This is a reason to use explicit requirements even when missing packages resolve automatically. Pin the data.table version being tested while allowing the resolver to choose dtplyr and compatible dependencies. Preparation makes packages available; the code still attaches them.

**Show:** Highlight the == version spec and the unpinned name in the same requirements.r array. ir supports == pins; do not replace this with an unsupported ad hoc notation.

**Author / capture:** Run this in a fresh session before data.table is loaded. A compatible environment addition is not a promise to replace an already loaded R namespace with another version. Local validation on 2026-09-17 failed: data.table 1.17.8 did not compile against R 4.6.1 (undeclared SETLENGTH/ATTRIB APIs). A separate R 4.5 attempt also failed during package installation with “worker failed to start.” The pin is retained as requested; captures/requirements and captures/requirements-r45 preserve the failed wire exchanges. Rehearse this example with a compatible, prepared runtime before presenting it live.

**Sources:** [requirements](https://github.com/t-kalinowski/mcp-console/blob/main/docs/REQUIREMENTS.md) · [ir](https://github.com/r-lib/ir/blob/main/README.md)

:::


## Prepare now; keep working in the same session {#live-requirements .deck-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="arg">requirements</span><span class="sig-name">)</span></div>

::::: {.two-col}

:::: {.left}

::: {.code-panel}

::: {.panel-label}

send

:::

```python
console.send(requirements={
    "r": ["data.table"],
})

console.send(r="coef(fit)")
```

:::

::::

:::: {.right}

::: {.big-statement}

New dependency.<br>Same objects.

:::

::::

:::::

::: {.takeaway}

Requirements-only calls work; compatible live additions preserve state.

:::

::: {.notes}

**Say:** Requirements can also be prepared without submitting a cell. Supported compatible additions keep the existing worker state, and accepted requirements remain retained for later calls and restart. This is additive environment management, not arbitrary replacement or removal of already loaded packages.

**Show:** Show a requirements-only call followed by inspection of the same fit. The new field has gained another use without another tool. Mention that some incompatible or failed activations require an explicit restart; do not imply arbitrary upgrades can preserve every live object.

**Sources:** [Requirements and environments](https://github.com/t-kalinowski/mcp-console/blob/main/docs/REQUIREMENTS.md)

:::



## Use the resolver, not another package manager {#ir .deck-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="arg">requirements</span><span class="sig-name">)</span></div>

::: {.big-statement}

R requirements → `ir`

:::

::: {.takeaway}

Prepare and reuse the package environment; leave session ownership to Console.

:::

::: {.small-note}

Installation and trust boundaries come later.

:::

::: {.notes}

**Say:** Console coordinates the environment but delegates R dependency work to ir. This matters for reuse and for keeping package preparation separate from the evaluator. We will return to concrete versus managed environments and the sandbox boundary after the language sequence.

**Show:** Use one large dependency arrow, not an installation walkthrough. It is a fast acknowledgement of the underlying tooling, while the signature stays the same.

**Sources:** [Requirements and environments](https://github.com/t-kalinowski/mcp-console/blob/main/docs/REQUIREMENTS.md)

:::



## Python can read the R workspace {#python-first .deck-slide .literal-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="arg">requirements</span><span class="arg new-arg">python</span><span class="sig-name">)</span></div>

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

Python cell

:::

<!-- cells:python-first -->
```python
console.send(python='print(len(r.d))')
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

Console response

:::

```{r}
#| eval: true
#| echo: false
#| comment: ""
console_protocol("python-first")
```

::::

:::::

::::::

::: {.takeaway}

`r.d` accesses the data frame already loaded in R.

:::

::: {.notes}

**Say:** Python works in the same session. Start with one expression: how many rows are in the R data frame? The r bridge exposes the existing object. Next we can use Python libraries on that data.

**Show:** This is an actual captured Python response.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::

## Use Python’s ML libraries on the same data {#python .deck-slide .code-dense}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="arg">requirements</span><span class="arg">python</span><span class="sig-name">)</span></div>
:::: {.code-panel}

::: {.panel-label}

LLM client → Console · Python cell

:::

<!-- cells:python-ml -->
```python
console.send(python="""
from sklearn.ensemble import RandomForestRegressor
from sklearn.model_selection import cross_val_score
import pandas as pd

frame = r.d
X = pd.get_dummies(frame[["temperature", "group"]])
model = RandomForestRegressor(random_state=42)
scores = cross_val_score(model, X, frame["response"],
                         cv=5, scoring="neg_mean_absolute_error")
print(f"CV MAE: {-scores.mean():.3f}")
""")
```

::::
::: {.takeaway}

For example: scikit-learn cross-validation, without exporting the R data.

:::

::: {.notes}

**Say:** A reason to switch languages is access to another ecosystem. Here Python reads the existing R frame and uses scikit-learn to evaluate a nonlinear model. This is not a claim that Python plots better, or that this small demonstration selects the best model.

**Show:** The new python argument is the only interface expansion. The meaningful difference is the library being used: RandomForestRegressor and cross_val_score. Show the literal result on the next slide.

**Sources:** [runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [requirements](https://github.com/t-kalinowski/mcp-console/blob/main/docs/REQUIREMENTS.md)

:::


## The Python cell returns a numerical result {#python-ml-result .deck-slide .literal-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="arg">requirements</span><span class="arg">python</span><span class="sig-name">)</span></div>
:::: {.result-panel}

::: {.panel-label}

Python output · reference or captured

:::

```{r}
#| label: python-ml-result-output
#| eval: true
#| echo: false
#| comment: ""
if (.console_mode == "mcp") {
  console_protocol("python-ml")
} else {
  cat(readLines("examples/python-ml-reference.txt"), sep = "\n")
}
```

::::
:::::: {.two-col}

::::: {.left}

::: {.ownership-card}

Same dataset

`frame = r.d`

:::

:::::

::::: {.right}

::: {.ownership-card}

Another library

`cross_val_score(...)`

:::

:::::

::::::


::: {.notes}

**Say:** The result is the five-fold cross-validation mean absolute error for the displayed code. It is an example computation, not a benchmark for Console or a claim about model quality. The data remain available in R and Python.

**Show:** Show the numerical print line, not a prose description of the score. The reference was computed locally in Python; switching the render parameter to mcp uses the real Console result.

**Author / capture:** Reference provenance and installed Python library versions are in examples/python-reference-provenance.json. Recompute or capture when changing code, data, or library versions.

**Sources:** [runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::


## Python plotting has the same return contract {#python-plots .deck-slide .literal-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="arg">requirements</span><span class="arg">python</span><span class="sig-name">)</span></div>
:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

LLM client → Console · optional Python plot

:::

<!-- cells:python-plot -->
```python
console.send(python="""
import matplotlib.pyplot as plt

plt.scatter(frame["temperature"],
            frame["response"])
plt.xlabel("Temperature")
plt.ylabel("Response")
""")
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

Matplotlib output

:::

```{r}
#| label: python-plot-output
#| eval: true
#| echo: false
#| comment: ""
#| out-width: "100%"
console_plot("python-plot")
```

::::

:::::

::::::
::: {.takeaway}

Plot in either language; image delivery is the common capability.

:::

::: {.notes}

**Say:** Plotting works here too, but plotting is not the reason we switched to Python. An open pyplot figure is returned as image content. This complements R graphics rather than ranking the two plotting systems.

**Show:** Use the actual Matplotlib result from the displayed code. The local reference and the captured image use the same data and call definition.

**Sources:** [runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::


## The controls come with it {#python-controls .deck-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="arg">requirements</span><span class="arg">python</span><span class="sig-name">)</span></div>

::: {.code-panel}

::: {.panel-label}

Independent call shapes

:::

```python
console.send(python="run_experiment()", timeout_ms=200)
console.send()
console.send(control="interrupt")
console.send(stdin="continue\n")
console.send(control="restart")
```

:::

::: {.takeaway}

Waits. Polls. Interrupts. Input. Debuggers. Restart.

:::

::: {.notes}

**Say:** These are independent example call shapes, not a script to execute consecutively. Python inherits the shared controls: bounded waits, polling, interruption, interactive input, supported debugger interaction, and explicit restart. There is no second Python-specific tool family to learn.

**Show:** Use a compact list of familiar calls rather than another control tutorial. Mark the panel as independent examples. The absence of a new signature field is the point.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [Send operation contract](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SEND_OPERATIONS.md)

:::



## Missing imports resolve in place {#python-resolution .deck-slide .literal-slide .code-dense}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="arg">requirements</span><span class="arg">python</span><span class="sig-name">)</span></div>

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

An ordinary import

:::

```python
console.send(python="""
import polars as pl
pl.from_pandas(r.d).group_by("group").len()
""")
```

::::

:::::

::::: {.right}

::: {.resolution-steps}

1. Python tries the import.
2. Normal lookup finds no module.
3. Console requests resolution.
4. `uv` prepares the environment.
5. The worker activates the result.
6. Python loads the module.

:::

:::::

::::::

::: {.notes}

**Say:** The final import finder runs after normal lookup cannot find the module, before a missing-module error escapes to the cell. It maps the import name to a distribution and calls the resolver. The server prepares the managed environment with uv; the worker activates the result, invalidates import caches, and continues loading the requested module. The rest of the cell then runs.

**Show:** Walk through the six steps in order. This managed path applies to supported missing imports; an explicitly selected Python environment uses its installed packages.

**Sources:** [Python import finder](https://github.com/t-kalinowski/mcp-console/blob/main/src/python/runtime.py) · [Requirements](https://github.com/t-kalinowski/mcp-console/blob/main/docs/REQUIREMENTS.md)

:::

## Versions and extras fit the existing field {#python-requirements .deck-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="arg">requirements</span><span class="arg">python</span><span class="sig-name">)</span></div>

::: {.code-panel}

::: {.panel-label}

send

:::

```python
console.send(
    python="import polars as pl; pl.__version__",
    requirements={"python": ["polars>=1,<2"]},
)
```

:::

::: {.takeaway}

Precise package metadata, without a new tool or a new workflow.

:::

::: {.notes}

**Say:** Requirements now carries Python distribution metadata: versions, extras, and markers within the accepted named-registry format. It can also correct an import-name inference. The model that understands the environment more precisely can use that precision through the same interface.

**Show:** Highlight the python entry inside requirements, not a new top-level field. This visually reinforces the difference between adding a capability to an existing contract and adding another tool.

**Sources:** [Requirements and environments](https://github.com/t-kalinowski/mcp-console/blob/main/docs/REQUIREMENTS.md) · [Canonical tool schema snapshot](https://github.com/t-kalinowski/mcp-console/blob/main/tests/snapshots/client_server/server/test_tools/initializes_and_lists_tools.yaml)

:::



## Both languages share one worker process {#shared-process .deck-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="arg">requirements</span><span class="arg">python</span><span class="sig-name">)</span></div>

<!-- inline-svg:assets/shared.svg -->
```{=html}
<div class="embedded-diagram">
<svg role="img" aria-label="R and Python are connected by reticulate inside one worker process; DuckDB also lives there." width="767pt" height="241pt"
 viewBox="0.00 0.00 766.80 240.80" xmlns="http://www.w3.org/2000/svg" xmlns:xlink="http://www.w3.org/1999/xlink">
<g id="diagram-4-graph0" class="graph" transform="scale(1 1) rotate(0) translate(14.4 226.4)">
<title>G</title>
<g id="diagram-4-clust1" class="cluster">
<title>cluster_worker</title>
<path fill="transparent" stroke="#167d76" stroke-width="2" d="M20,-8C20,-8 718,-8 718,-8 724,-8 730,-14 730,-20 730,-20 730,-192 730,-192 730,-198 724,-204 718,-204 718,-204 20,-204 20,-204 14,-204 8,-198 8,-192 8,-192 8,-20 8,-20 8,-14 14,-8 20,-8"/>
<text text-anchor="middle" x="369" y="-184" font-family="Arial" font-size="20.00" fill="#263d46">ONE WORKER PROCESS</text>
</g>
<!-- r -->
<g id="diagram-4-node1" class="node">
<title>r</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M202,-105C202,-105 44,-105 44,-105 38,-105 32,-99 32,-93 32,-93 32,-51 32,-51 32,-45 38,-39 44,-39 44,-39 202,-39 202,-39 208,-39 214,-45 214,-51 214,-51 214,-93 214,-93 214,-99 208,-105 202,-105"/>
<text text-anchor="middle" x="123" y="-77.6" font-family="Arial" font-size="18.00" fill="#142d39">R</text>
<text text-anchor="middle" x="123" y="-57.6" font-family="Arial" font-size="18.00" fill="#142d39">global environment</text>
</g>
<!-- bridge -->
<g id="diagram-4-node2" class="node">
<title>bridge</title>
<path fill="transparent" stroke="transparent" stroke-width="1.5" d="M478.83,-52C478.83,-52 419.17,-52 419.17,-52 415.83,-52 412.5,-48.67 412.5,-45.33 412.5,-45.33 412.5,-38.67 412.5,-38.67 412.5,-35.33 415.83,-32 419.17,-32 419.17,-32 478.83,-32 478.83,-32 482.17,-32 485.5,-35.33 485.5,-38.67 485.5,-38.67 485.5,-45.33 485.5,-45.33 485.5,-48.67 482.17,-52 478.83,-52"/>
<text text-anchor="middle" x="449" y="-37.6" font-family="Arial" font-size="18.00" fill="#142d39">reticulate</text>
</g>
<!-- r&#45;&gt;bridge -->
<g id="diagram-4-edge1" class="edge">
<title>r&#45;&gt;bridge</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M222.02,-62.92C282.99,-57.28 358.37,-50.3 404.81,-46"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="221.48,-60.34 214.25,-63.64 221.96,-65.57 221.48,-60.34"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="405.06,-48.61 412.29,-45.31 404.58,-43.38 405.06,-48.61"/>
</g>
<!-- db -->
<g id="diagram-4-node4" class="node">
<title>db</title>
<path fill="#d9eee9" stroke="#167d76" stroke-width="1.5" d="M535,-150C535,-150 363,-150 363,-150 357,-150 351,-144 351,-138 351,-138 351,-96 351,-96 351,-90 357,-84 363,-84 363,-84 535,-84 535,-84 541,-84 547,-90 547,-96 547,-96 547,-138 547,-138 547,-144 541,-150 535,-150"/>
<text text-anchor="middle" x="449" y="-122.6" font-family="Arial" font-size="18.00" fill="#142d39">DuckDB</text>
<text text-anchor="middle" x="449" y="-102.6" font-family="Arial" font-size="18.00" fill="#142d39">connection + catalog</text>
</g>
<!-- r&#45;&gt;db -->
<g id="diagram-4-edge3" class="edge">
<title>r&#45;&gt;db</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M214.25,-84.54C254.35,-90.1 301.79,-96.69 343.43,-102.48"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="343.07,-105.08 350.86,-103.51 343.79,-99.88 343.07,-105.08"/>
<text text-anchor="middle" x="282.5" y="-102.8" font-family="Arial" font-size="14.00" fill="#536b72">live R relations</text>
</g>
<!-- py -->
<g id="diagram-4-node3" class="node">
<title>py</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M694,-98C694,-98 608,-98 608,-98 602,-98 596,-92 596,-86 596,-86 596,-44 596,-44 596,-38 602,-32 608,-32 608,-32 694,-32 694,-32 700,-32 706,-38 706,-44 706,-44 706,-86 706,-86 706,-92 700,-98 694,-98"/>
<text text-anchor="middle" x="651" y="-70.6" font-family="Arial" font-size="18.00" fill="#142d39">Python</text>
<text text-anchor="middle" x="651" y="-50.6" font-family="Arial" font-size="18.00" fill="#142d39">__main__</text>
</g>
<!-- bridge&#45;&gt;py -->
<g id="diagram-4-edge2" class="edge">
<title>bridge&#45;&gt;py</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M493.54,-47.01C521.2,-50.19 557.43,-54.35 588.05,-57.88"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="493.51,-44.36 485.76,-46.11 492.91,-49.58 493.51,-44.36"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="587.96,-60.51 595.71,-58.76 588.56,-55.29 587.96,-60.51"/>
</g>
</g>
</svg>
</div>
```
<!-- /inline-svg -->

::: {.takeaway}

Shared lifetime; explicit object bridges; no intermediate files required.

:::

::: {.notes}

**Say:** In the combined runtime, R hosts Python through reticulate and DuckDB is in the same worker process. R globals are available through r in Python; Python globals through py in R. Conversions follow reticulate’s rules. Do not claim arbitrary zero-copy conversion or that the host R chat session is this worker.

**Show:** Use a single visible process boundary. The R/Python bridge is inside it. This is an overview of the current combined-runtime topology, not a claim that the planned Python-only mode is already this same implementation.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [Implemented architecture](https://github.com/t-kalinowski/mcp-console/blob/main/docs/ARCHITECTURE.md)

:::



## Switch languages; reuse the objects {#object-bridge .deck-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="arg">requirements</span><span class="arg">python</span><span class="sig-name">)</span></div>

::: {.code-panel}

::: {.panel-label}

send

:::

```python
console.send(python="""
frame = r.d
group_counts = frame.groupby("group").size()
""")

console.send(r="py$group_counts")
```

:::



::: {.notes}

**Say:** Now make the object bridge concrete. A Python cell works on the R data frame and stores a Python result; the next R cell reads that Python object through py. The session stays live throughout. File export/import is not part of the normal handoff, though conversions can still allocate.

**Show:** Show the two calls stacked in one panel. Reuse variable names from the preceding slides so the audience can follow state without rereading a new scenario.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [Project README](https://github.com/t-kalinowski/mcp-console/blob/main/README.md)

:::



## And SQL {#sql .deck-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="arg">requirements</span><span class="arg">python</span><span class="arg new-arg">sql</span><span class="sig-name">)</span></div>

::: {.code-panel}

::: {.panel-label}

send

:::

```python
console.send(sql="""
SELECT "group", COUNT(*) AS n
FROM d
GROUP BY "group"
""")
```

:::

::: {.takeaway}

DuckDB by default. Persistent connection and catalog.

:::

::: {.notes}

**Say:** Add sql, the last code-language field. SQL has its own cell input and a persistent managed DuckDB backend by default. The d relation here is the live R data frame already created; an unqualified relation name can discover it. A DuckDB table or view of the same name takes precedence.

**Show:** Reveal sql in the signature. The rest of the call/result composition remains unchanged. Again, r, python, and sql are alternatives per call, not three simultaneous code payloads.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [Canonical tool schema snapshot](https://github.com/t-kalinowski/mcp-console/blob/main/tests/snapshots/client_server/server/test_tools/initializes_and_lists_tools.yaml)

:::



## SQL can operate on live R data {#sql-r-data .deck-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="arg">requirements</span><span class="arg">python</span><span class="arg">sql</span><span class="sig-name">)</span></div>

::: {.code-panel}

::: {.panel-label}

send

:::

```python
console.send(r="d$.residual <- residuals(fit)")

console.send(sql="""
SELECT "group", AVG(ABS(".residual")) AS error
FROM d
GROUP BY "group"
""")
```

:::



::: {.notes}

**Say:** An R model result becomes a column, and SQL immediately summarizes it through managed DuckDB relation discovery. This is the point of one interactive workspace: each language builds on live work already done. The default managed backend discovers R frames; Python visibility is connection-dependent and needs the explicit setup shown next.

**Show:** Show a one-line R update followed by the SQL query. Keep the claim precise: this is live R relation discovery on the managed backend, not universal implicit visibility into every object in every runtime.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [Canonical tool schema snapshot](https://github.com/t-kalinowski/mcp-console/blob/main/tests/snapshots/client_server/server/test_tools/initializes_and_lists_tools.yaml)

:::



## Python relations use their connection {#sql-python-data .deck-slide .code-dense}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="arg">requirements</span><span class="arg">python</span><span class="arg">sql</span><span class="sig-name">)</span></div>

::: {.code-panel}

::: {.panel-label}

send

:::

```python
console.send(python="""
import duckdb
con = duckdb.connect()
con.register("measurements", r.d)
console_sql_connection(con)
""")

console.send(sql="SELECT COUNT(*) FROM measurements")
```

:::

::: {.takeaway}

Connection-local registrations stay with the Python connection.

:::

::: {.notes}

**Say:** For Python-side relations, select a Python DuckDB connection and register the frame there. Later SQL cells use that exact connection. This avoids the inaccurate implication that managed R-backed DuckDB automatically sees Python globals. Another route is to bind a Python frame to an R name before querying it through the managed backend.

**Show:** Keep registration visible: it is useful explicit runtime capability, not a new Console tool. The example uses the existing R frame via Python only to preserve the narrative; the frame could just as well have been created in Python.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [Canonical tool schema snapshot](https://github.com/t-kalinowski/mcp-console/blob/main/tests/snapshots/client_server/server/test_tools/initializes_and_lists_tools.yaml)

:::



## Select the database from either language {#sql-connections .deck-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="arg">requirements</span><span class="arg">python</span><span class="arg">sql</span><span class="sig-name">)</span></div>

::::: {.two-col}

:::: {.left}

::: {.code-panel}

::: {.panel-label}

R cell · DBI

:::

```r
connection <- DBI::dbConnect(
  RSQLite::SQLite(), ":memory:"
)
console_sql_connection(connection)
```

:::

::::

:::: {.right}

::: {.code-panel}

::: {.panel-label}

Python cell · DB-API

:::

```python
import sqlite3
connection = sqlite3.connect(
    ":memory:"
)
console_sql_connection(connection)
```

:::

::::

:::::

::: {.takeaway}

The selected driver owns dialect, transactions, and relation visibility.

:::

::: {.notes}

**Say:** DuckDB is the default, not a limitation of the SQL cell interface. R can select a DBI connection and Python can select a DB-API connection. The connection remains owned by its runtime. SQL semantics, transactions, and supported statements belong to the selected driver; these examples are alternatives, not shared cross-runtime connection objects.

**Show:** Use parallel code panels and one shared takeaway. The function console_sql_connection() belongs to runtime code, demonstrating how much functionality can be exposed without changing send.

**Sources:** [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [Canonical tool schema snapshot](https://github.com/t-kalinowski/mcp-console/blob/main/tests/snapshots/client_server/server/test_tools/initializes_and_lists_tools.yaml)

:::



## And DuckDB extensions {#duckdb-requirements .deck-slide}
::: {.kicker}

Build the interface

:::

<div class="signature" aria-label="Fields introduced so far"><span class="sig-name">console.send(</span><span class="arg">r</span><span class="arg">timeout_ms</span><span class="arg">control</span><span class="arg">stdin</span><span class="arg">requirements</span><span class="arg">python</span><span class="arg">sql</span><span class="sig-name">)</span></div>

::: {.code-panel}

::: {.panel-label}

send

:::

```python
console.send(
    sql="LOAD spatial; SELECT ST_Point(1, 2)",
    requirements={"duckdb": ["spatial"]},
)
```

:::

::: {.takeaway}

Prepare explicitly—or resolve on demand—then load inside the worker.

:::

::: {.notes}

**Say:** Dependency preparation also covers DuckDB extensions. The explicit example is supported for the managed backend; preparation uses DuckDB’s install path outside the worker sandbox, while loading happens inside the runtime. For presentation day, include the intended on-demand extension path as well, without inventing a new top-level field.

**Show:** Show one extension requirement beside a query. Reset to the managed DuckDB connection before actually running this example after custom-connection slides. The automatic on-demand claim is a target-day assumption and must be verified before a live capture.

**Author check before presenting:** Target-day assumption: automatic DuckDB extension discovery is not documented as implemented; current SQL does not trigger package discovery. Explicit requirements.duckdb is implemented.

**Sources:** [Requirements and environments](https://github.com/t-kalinowski/mcp-console/blob/main/docs/REQUIREMENTS.md) · [Canonical tool schema snapshot](https://github.com/t-kalinowski/mcp-console/blob/main/tests/snapshots/client_server/server/test_tools/initializes_and_lists_tools.yaml)

:::



## That is the surface {#full-signature .deck-slide .signature-slide}
::: {.kicker}

The complete contract

:::

::: {.full-contract}

<pre class="signature-display">console.send(
  <span class="syntax-accent">r</span>=…, <span class="syntax-accent">python</span>=…, <span class="syntax-accent">sql</span>=…,   <span class="syntax-comment"># at most one code field</span>
  timeout_ms=…,
  control=…,
  stdin=…,
  requirements=…,
)</pre>

:::

::: {.takeaway}

Seven top-level fields. All optional. One tool.

:::

::: {.small-note}

Poll, input, control, and prepare-only calls omit code.

:::

::: {.notes}

**Say:** Now reveal the complete top-level surface. All fields are optional, with compatibility rules on combinations; at most one code-language field is supplied. The fields are not seven separate tools. Much of the capability just demonstrated lives in native runtime operations, so the schema does not need a method for every package, plot, connection, or debugger action.

**Show:** Drop the incremental signature rail and replace it with a single large schematic signature. Be explicit that this is named-argument shorthand for the MCP object schema, not a literal positional API declaration. Let the size contrast with the capabilities already demonstrated.

**Sources:** [Canonical tool schema snapshot](https://github.com/t-kalinowski/mcp-console/blob/main/tests/snapshots/client_server/server/test_tools/initializes_and_lists_tools.yaml) · [Send operation contract](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SEND_OPERATIONS.md)

:::



## Start simply; opt into more control {#model-capability .deck-slide}
::: {.kicker}

The complete contract

:::

:::::: {.two-col}

::::: {.left}

::: {.ownership-card}

Start with ordinary code

`library(package)`

`import package`

Console prepares missing dependencies.

:::

:::::

::::: {.right}

::: {.ownership-card}

Discover more as needed

Version pins and environments

Polling and interruption

Debugging and language bridges

:::

:::::

::::::

::: {.notes}

**Say:** Models can hesitate to install packages, and simpler models may stay with basic code. We support that path: ordinary package use can ask the runtime to prepare what is missing. As a model becomes more capable, it can discover and use the longer tail of controls—requirements, polling, debugging, and language bridges—through the same tool. This is a design motivation, rather than a measured comparison of model performance.

**Sources:** [Requirements](https://github.com/t-kalinowski/mcp-console/blob/main/docs/REQUIREMENTS.md)

:::

## An interaction journal, and the actual output {#recording .deck-slide .literal-slide}
::: {.kicker}

Records and artifacts

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

Files under one session directory

:::

```text
session/
├── internal/
│   └── events.jsonl
├── outputs/
│   └── call-000003.log
├── artifacts/
│   └── *.png
├── transcript.md
└── transcript.qmd
```

::::

:::::

::::: {.right}

:::: {.result-panel}

::: {.panel-label}

Raw cell log · fit output

:::

```{r}
#| eval: true
#| echo: false
#| comment: ""
console_protocol("excerpts/recorded-fit")
```

::::

:::::

::::::

::: {.notes}

**Say:** The indentation shows the actual layout. Raw cell logs live in outputs, PNGs in artifacts, and the two transcripts are siblings of those directories. The internal journal preserves Console requests—including harness metadata—and assembled results, plus recording lifecycle events.

**Show:** The right panel reads the captured fit log. The next slide shows a journal event with request metadata. Full transport traffic, including initialization, is separately captured by our collector in captures/wire.jsonl.

**Sources:** [Architecture](https://github.com/t-kalinowski/mcp-console/blob/main/docs/ARCHITECTURE.md)

:::

## Preserve the request and its metadata {#recording-event .deck-slide .literal-slide .code-dense}
::: {.kicker}

Records and artifacts

:::

:::: {.code-panel}

::: {.panel-label .path-label}

internal/events.jsonl · actual event rendered as YAML

:::

```{r}
#| eval: true
#| echo: false
#| results: asis
console_source("event-yaml", "yaml")
```

::::

::: {.notes}

**Say:** This is a complete tool_call journal event from the capture, converted from JSON to YAML. The harness supplied progressToken and an example namespaced metadata object; both survived in request._meta. These are demonstration metadata, not a claim that every harness uses these keys. The event also records the request ID and Console sequence.

**Show:** Call arguments and metadata stay together. The journal records send calls and results; our separate wire.jsonl contains the full JSON-RPC exchange.

**Sources:** [Recorder](https://github.com/t-kalinowski/mcp-console/blob/main/src/transcript.rs)

:::

## The Markdown contains code and results {#markdown-transcript .deck-slide .literal-slide .transcript-slide}
::: {.kicker}

Records and artifacts

:::

:::: {.code-panel}

::: {.panel-label}

transcript.md · one complete call and result

:::

```{r}
#| label: markdown-excerpt
#| eval: true
#| echo: false
#| comment: ""
#| results: asis
console_source("markdown", "markdown")
```

::::

::: {.takeaway}

A readable call ledger: source, returned text, and image links.

:::

::: {.notes}

**Say:** This is the kind of content to inspect in transcript.md: the submitted R cell followed by literal returned text. Real transcripts additionally retain call options, errors, input interactions, and image references. Polls remain separate calls rather than being silently regrouped as one evaluation.

**Show:** Use a concrete excerpt from the same analysis. Read the actual call heading, R source fence, result heading, and returned coefficients.

**Author / capture:** This is a contiguous, literal excerpt of captures/session-records/transcript.md: one complete coefficient call and its result. The initial session metadata and other calls are outside the excerpt. No identifiers, values, or paths were normalized; line ranges are in captures/excerpts/provenance.json.

**Sources:** [architecture](https://github.com/t-kalinowski/mcp-console/blob/main/docs/ARCHITECTURE.md)

:::


## The Quarto source is editable and executable {#quarto-transcript .deck-slide .literal-slide .code-dense}
::: {.kicker}

Records and artifacts

:::

:::::: {.code-panel}

::: {.panel-label}

transcript.qmd · actual front matter and model cell

:::

::::: {.two-col}

::: {.left}

```{r}
#| label: quarto-frontmatter
#| eval: true
#| echo: false
#| comment: ""
#| results: asis
header <- console_capture_text("excerpts/quarto-header")
header <- sub("root.dir: [^\n]+", "root.dir: <project>", header)
cat("````yaml\n", header, "\n````\n", sep = "")
```

:::

::: {.right}

```{r}
#| label: quarto-cell
#| eval: true
#| echo: false
#| comment: ""
#| results: asis
console_source("quarto-cell", "markdown")
```

:::

:::::

::::::



::: {.notes}

**Say:** The source projection contains executable chunks rather than escaped code in a log. Rendering runs those cells again in a new Quarto environment. It does not replay stdin, interrupts, restarts, or prior outputs, and the required data and SQL connection still need to exist.

**Show:** Show literal Quarto chunk delimiters. The left panel contains the complete generated front matter. The right panel is the unchanged model cell; the preceding warmup cell and later cells are outside the excerpt.

**Author / capture:** These are literal source slices from the captured document. The long root.dir path is shortened to <project> only in this slide; the original is preserved in captures/excerpts/quarto-header.txt. Dependencies reflect the generated front matter of this build and do not include every inferred package. Local render occurs outside the Console worker sandbox. Remote projections default eval:false.

**Sources:** [architecture](https://github.com/t-kalinowski/mcp-console/blob/main/docs/ARCHITECTURE.md)

:::


## The user configures the boundary {#config .deck-slide .config-slide}
::: {.kicker}

Sandbox · configure the boundary

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label .path-label}

.agents/console/config.yaml

:::

```yaml
extends: ":workspace"
sandbox:
  provider: native
  network: restricted
```

::::

:::::

::::: {.right}

::: {.ownership-card}

Filesystem access

Networking and proxy policy

Environment and execution target

:::

:::::

::::::
::: {.takeaway}

Trusted launch input. One captured configuration per server session.

:::

::: {.notes}

**Say:** The agent-facing send interface is not a policy editor. The user chooses the execution conditions in a project YAML file. Console reads it from the launch directory, applies explicit command-line overrides, and captures the result before workers start.

**Show:** Start with a short complete file and name the categories we will expand. Do not show proposed top-level profile or environment keys as implemented schema.

**Sources:** [config](https://github.com/t-kalinowski/mcp-console/blob/main/docs/CONFIGURATION.md) · [sandbox](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md) · [settings](https://github.com/t-kalinowski/mcp-console/blob/main/src/settings.rs)

:::


## Keep the sandbox behind a replaceable boundary {#external-runner .deck-slide}
::: {.kicker}

Sandbox · configure the boundary

:::

:::::: {.two-col}

::::: {.left}

::: {.ownership-card}

Console

Reads the project configuration.

Starts and communicates with the worker.

:::

:::::

::::: {.right}

::: {.ownership-card}

Sandbox implementation

The native runner is a standalone binary.

It applies the OS restrictions and launches the workload.

:::

:::::

::::::

::: {.takeaway}

The execution boundary can change while the Console interface stays the same.

:::

::: {.notes}

**Say:** Native means the standalone mcp-console-sandbox executable that applies the platform’s restrictions. Console supplies policy and launches the workload through that boundary. Keeping it separate allows the sandbox implementation to evolve independently. The implemented alternative is the compute provider used by Docker Sandbox, shown later. A different native executable must implement the runner protocol; this is not a promise that an arbitrary command is a drop-in replacement.

The native external-sandbox filesystem mode means an existing outer environment supplies enforcement. It does not create that environment. This detail belongs in configuration reference material rather than a separate slide.

**Sources:** [Sandbox configuration](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md) · [Runner protocol](https://github.com/t-kalinowski/codex/blob/2d0ad797210de821c07d1f18e4f1ffdcf06589cb/codex-rs/mcp-console-sandbox/PROTOCOL.md)

:::

## Control the workload’s environment {#environment-choice .deck-slide .config-slide .code-dense}
::: {.kicker}

Sandbox · configure the boundary

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label .path-label}

.agents/console/config.yaml

:::

```yaml
extends: ":read-only"
sandbox:
  inherit_environment: false
  environment:
    ANALYSIS_MODE: interactive
    OMP_NUM_THREADS: "4"
    MPLBACKEND: Agg
```

::::

:::::

::::: {.right}

::: {.ownership-card}

Do not inherit the ordinary environment.

Set explicit workload variables.

Selected runtime bindings remain Console-owned.

:::

:::::

::::::


::: {.notes}

**Say:** inherit_environment controls ordinary workload variables. It does not erase Console’s selected R/Python runtime settings or configure the trusted dependency-preparation process. Values that look numeric must still be strings.

**Show:** Walk through the visible YAML from top to bottom. This is a complete alternative configuration, not a patch to concatenate with the preceding example.

**Author / capture:** This limits inherited variables, not filesystem access to secrets. Local resolver settings come from the trusted server launch environment; SSH runtime exceptions are documented separately.

**Sources:** [sandbox](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::


## Start from the read-only baseline {#default-sandbox .deck-slide .config-slide .code-dense}
::: {.kicker}

Sandbox · configure the boundary

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label .path-label}

.agents/console/config.yaml

:::

```yaml
extends: ":read-only"
```

::::

:::::

::::: {.right}

::: {.ownership-card}

Host files: readable.

Private scratch: writable.

Direct worker network: restricted.

:::

:::::

::::::


::: {.notes}

**Say:** Read-only means ordinary host writes are restricted, not that the worker cannot read the machine or write its private temporary files. Native sandboxing is the default on supported hosts. This is a process boundary around evaluated code and descendants.

**Show:** Walk through the visible YAML from top to bottom. This is a complete alternative configuration, not a patch to concatenate with the preceding example.

**Sources:** [sandbox](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::


## Permit project edits, protect metadata by default {#permission-grants .deck-slide .config-slide .code-dense}
::: {.kicker}

Sandbox · configure the boundary

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label .path-label}

.agents/console/config.yaml

:::

```yaml
extends: ":workspace"
```

::::

:::::

::::: {.right}

::: {.ownership-card}

Workspace: writable.

.git · .agents · .codex · .claude: read-only defaults.

Network: still restricted.

:::

:::::

::::::


::: {.notes}

**Say:** The workspace profile grants project edits while protecting common repository and agent metadata by default. The fixed launch workspace is the permission root; changing the R working directory does not move it. Explicit native rules can override defaults.

**Show:** Walk through the visible YAML from top to bottom. This is a complete alternative configuration, not a patch to concatenate with the preceding example.

**Sources:** [sandbox](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::


## Grant one persistent output directory {#write-output .deck-slide .config-slide .code-dense}
::: {.kicker}

Sandbox · configure the boundary

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label .path-label}

.agents/console/config.yaml

:::

```yaml
extends: ":read-only"
sandbox:
  filesystem:
    entries:
      - path: {type: path, path: ./output}
        access: write
```

::::

:::::

::::: {.right}

::: {.ownership-card}

Only ./output gains writes.

The rest retains read-only access.

Relative paths use the launch directory.

:::

:::::

::::::
::: {.takeaway}

A narrow write grant; no implicit network grant.

:::

::: {.notes}

**Say:** Grant a specific directory rather than the entire project when the task only needs to save results. These are literal native path objects. Pre-create the directory for a portable Linux demonstration; Console does not widen the grant to its parent.

**Show:** Walk through the visible YAML from top to bottom. This is a complete alternative configuration, not a patch to concatenate with the preceding example.

**Author / capture:** Create output before launching on Linux. Existing file-root and missing-path behavior differs by backend; do not promise identical behavior for every path kind.

**Sources:** [sandbox](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::


## Read data, write results, hide secrets {#private-inputs .deck-slide .config-slide .code-dense}
::: {.kicker}

Sandbox · configure the boundary

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label .path-label}

.agents/console/config.yaml

:::

```yaml
extends: ":workspace"
sandbox:
  filesystem:
    entries:
      - path: {type: path, path: ./data}
        access: read
      - path: {type: path, path: ./secrets}
        access: deny
      - path: {type: path, path: ./.env}
        access: deny
```

::::

:::::

::::: {.right}

::: {.ownership-card}

./data: read-only.

./secrets and ./.env: denied.

Other workspace paths retain their baseline.

:::

:::::

::::::
::: {.takeaway}

Different access modes express different capabilities.

:::

::: {.notes}

**Say:** Now distinguish a read grant from a denial. Making data read-only prevents accidental modification; deny is the separate choice that prevents reading that path. Native specificity, symlink handling, and platform enforcement still matter.

**Show:** Walk through the visible YAML from top to bottom. This is a complete alternative configuration, not a patch to concatenate with the preceding example.

**Author / capture:** Use pre-existing fixture paths. Native specificity is not array order. Linux has documented limits for nested deny/read reopenings; this simple example does not establish arbitrary cross-platform ACL equivalence.

**Sources:** [sandbox](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::


## Enable direct networking explicitly {#network-direct .deck-slide .config-slide .code-dense}
::: {.kicker}

Sandbox · configure the boundary

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label .path-label}

.agents/console/config.yaml

:::

```yaml
extends: ":read-only"
sandbox:
  network: enabled
```

::::

:::::

::::: {.right}

::: {.ownership-card}

Outbound access is a separate knob.

Filesystem writes remain restricted.

No domain allowlist in this example.

:::

:::::

::::::
::: {.takeaway}

Filesystem access and network access are separate controls.

:::

::: {.notes}

**Say:** network: enabled is the broad direct-network choice. It does not grant filesystem writes. The following slide uses a managed proxy to make a narrower network choice instead.

**Show:** Walk through the visible YAML from top to bottom. This is a complete alternative configuration, not a patch to concatenate with the preceding example.

**Sources:** [sandbox](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::


## Allow selected destinations through a proxy {#network-proxy .deck-slide .config-slide .code-dense}
::: {.kicker}

Sandbox · configure the boundary

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label .path-label}

.agents/console/config.yaml

:::

```yaml
extends: ":workspace"
sandbox:
  network: restricted
  proxy:
    enabled: true
    enableSocks5: true
    enableSocks5Udp: false
    allowUpstreamProxy: false
    dangerouslyAllowAllUnixSockets: false
    allowLocalBinding: false
    mode: full
    domains:
      warehouse.example.com: allow
      blocked.example.com: deny
```

::::

:::::

::::: {.right}

::: {.ownership-card}

Route worker traffic through the managed proxy.

Allow the corporate warehouse host.

Use its HTTPS database API, or a SOCKS-aware database connection.

:::

:::::

::::::
::: {.takeaway}

The client still supplies its usual database credentials.

:::

::: {.notes}

**Say:** For example, a corporate warehouse can expose an HTTPS query endpoint at the allowed host. A database using another TCP protocol needs a SOCKS-aware client or an explicitly configured SOCKS tunnel; a domain allowlist does not rewrite arbitrary database drivers. Network reachability and database authentication remain separate. For a database on a private IP, the pinned proxy requires a literal allowed IP matching the destination (or the broader local-binding exception); a hostname allowlist alone does not grant private-address access. The placeholder host is illustrative and was not contacted.

 This complete proxy configuration shows both the allowlist and the surrounding controls. The runner owns proxy routing, host normalization, and enforcement. A proxy object must include the required scalar fields; Console does not materialize omitted defaults for it.

**Show:** Walk through the visible YAML from top to bottom. This is a complete alternative configuration, not a patch to concatenate with the preceding example.

**Author / capture:** Example domains are placeholders, not actual services. All seven required scalar proxy fields are present. mode: full does not erase the domain allowlist. Review PROTOCOL.md for mode, upstream proxy, UDP and local-binding semantics.

**Sources:** [sandbox](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::


## Develop a local Shiny app {#network-listener .deck-slide .config-slide .code-dense}
::: {.kicker}

Sandbox · configure the boundary

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label .path-label}

.agents/console/config.yaml

:::

```yaml
extends: ":workspace"
sandbox:
  network: restricted
  proxy:
    enabled: true
    enableSocks5: true
    enableSocks5Udp: false
    allowUpstreamProxy: false
    dangerouslyAllowAllUnixSockets: false
    allowLocalBinding: true
    mode: full
    domains:
      api.example.com: allow
```

::::

:::::

::::: {.right}

::: {.ownership-card}

allowLocalBinding: true

Run Shiny on the loopback interface.

Open `http://127.0.0.1:8000` in your browser.

:::

:::::

::::::
::: {.takeaway}

`shiny::runApp("app", host = "127.0.0.1", port = 8000)`

:::

::: {.notes}

**Say:** This configuration supports local Shiny development: the worker can bind the loopback port while outbound requests use the managed proxy. Start an app at app/ with shiny::runApp and open its URL locally. A long-running app uses the wait, poll, and interrupt contract introduced earlier.

 An interactive local server can need a binding exception. The policy exposes that choice separately from the domain allowlist. Do not conflate permission to bind with permission to publish a remote service, create a tunnel, or expose every Unix socket.

**Show:** Walk through the visible YAML from top to bottom. This is a complete alternative configuration, not a patch to concatenate with the preceding example.

**Author / capture:** Validate the exact native behavior with the application used for the demo. This slide does not promise a Shiny viewer or automatic port forwarding. Keep all required proxy fields when changing a single flag.

**Sources:** [sandbox](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::


## Resolution crosses a different trust boundary {#resolver-boundary .deck-slide .config-slide .compact-diagram}
::: {.kicker}

Sandbox · configure the boundary

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label}

Worker stays restricted

:::

```yaml
extends: ":read-only"
sandbox:
  network: restricted
```

::::

:::::

::::: {.right}

:::: {.code-panel}

::: {.panel-label}

Model requests a dependency

:::

```python
console.send(
    r="library(dtplyr)",
    requirements={"r": ["dtplyr"]},
)
```

::::

:::::

::::::
<!-- inline-svg:assets/resolution.svg -->
```{=html}
<div class="embedded-diagram">
<svg role="img" aria-label="The worker requests package preparation outside its sandbox; trusted ir and uv resolvers prepare candidates for activation." width="958pt" height="285pt"
 viewBox="0.00 0.00 957.80 284.80" xmlns="http://www.w3.org/2000/svg" xmlns:xlink="http://www.w3.org/1999/xlink">
<g id="diagram-7-graph0" class="graph" transform="scale(1 1) rotate(0) translate(14.4 270.4)">
<title>G</title>
<g id="diagram-7-clust1" class="cluster">
<title>cluster_worker</title>
<path fill="transparent" stroke="#167d76" d="M20,-8C20,-8 276,-8 276,-8 282,-8 288,-14 288,-20 288,-20 288,-236 288,-236 288,-242 282,-248 276,-248 276,-248 20,-248 20,-248 14,-248 8,-242 8,-236 8,-236 8,-20 8,-20 8,-14 14,-8 20,-8"/>
<text text-anchor="middle" x="148" y="-229.6" font-family="Arial" font-size="18.00" fill="#263d46">WORKER SANDBOX</text>
</g>
<!-- load -->
<g id="diagram-7-node1" class="node">
<title>load</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M233,-196C233,-196 63,-196 63,-196 57,-196 51,-190 51,-184 51,-184 51,-142 51,-142 51,-136 57,-130 63,-130 63,-130 233,-130 233,-130 239,-130 245,-136 245,-142 245,-142 245,-184 245,-184 245,-190 239,-196 233,-196"/>
<text text-anchor="middle" x="148" y="-168.6" font-family="Arial" font-size="18.00" fill="#142d39">library() / import</text>
<text text-anchor="middle" x="148" y="-148.6" font-family="Arial" font-size="18.00" fill="#142d39">missing dependency</text>
</g>
<!-- server -->
<g id="diagram-7-node3" class="node">
<title>server</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M594,-191C594,-191 378,-191 378,-191 372,-191 366,-185 366,-179 366,-179 366,-137 366,-137 366,-131 372,-125 378,-125 378,-125 594,-125 594,-125 600,-125 606,-131 606,-137 606,-137 606,-179 606,-179 606,-185 600,-191 594,-191"/>
<text text-anchor="middle" x="486" y="-163.6" font-family="Arial" font-size="18.00" fill="#142d39">Trusted preparation owner</text>
<text text-anchor="middle" x="486" y="-143.6" font-family="Arial" font-size="18.00" fill="#142d39">validate + retain</text>
</g>
<!-- load&#45;&gt;server -->
<g id="diagram-7-edge1" class="edge">
<title>load&#45;&gt;server</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M245.28,-161.57C280.41,-161.04 320.68,-160.45 358.06,-159.89"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="358.53,-162.51 365.99,-159.77 358.45,-157.26 358.53,-162.51"/>
<text text-anchor="middle" x="319" y="-163.8" font-family="Arial" font-size="14.00" fill="#536b72">request</text>
</g>
<!-- live -->
<g id="diagram-7-node2" class="node">
<title>live</title>
<path fill="#d9eee9" stroke="#167d76" stroke-width="1.5" d="M252,-98C252,-98 44,-98 44,-98 38,-98 32,-92 32,-86 32,-86 32,-44 32,-44 32,-38 38,-32 44,-32 44,-32 252,-32 252,-32 258,-32 264,-38 264,-44 264,-44 264,-86 264,-86 264,-92 258,-98 252,-98"/>
<text text-anchor="middle" x="148" y="-70.6" font-family="Arial" font-size="18.00" fill="#142d39">Same live runtime</text>
<text text-anchor="middle" x="148" y="-50.6" font-family="Arial" font-size="18.00" fill="#142d39">resume original operation</text>
</g>
<!-- resolver -->
<g id="diagram-7-node4" class="node">
<title>resolver</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M917,-134C917,-134 713,-134 713,-134 707,-134 701,-128 701,-122 701,-122 701,-80 701,-80 701,-74 707,-68 713,-68 713,-68 917,-68 917,-68 923,-68 929,-74 929,-80 929,-80 929,-122 929,-122 929,-128 923,-134 917,-134"/>
<text text-anchor="middle" x="815" y="-106.6" font-family="Arial" font-size="18.00" fill="#142d39">ir / uv / DuckDB</text>
<text text-anchor="middle" x="815" y="-86.6" font-family="Arial" font-size="18.00" fill="#142d39">prepare outside sandbox</text>
</g>
<!-- server&#45;&gt;resolver -->
<g id="diagram-7-edge2" class="edge">
<title>server&#45;&gt;resolver</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M606.44,-137.18C634.75,-132.25 664.95,-126.98 693.31,-122.04"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="693.93,-124.59 700.87,-120.72 693.03,-119.42 693.93,-124.59"/>
<text text-anchor="middle" x="653.5" y="-135.8" font-family="Arial" font-size="14.00" fill="#536b72">prepare</text>
</g>
<!-- resolver&#45;&gt;live -->
<g id="diagram-7-edge3" class="edge">
<title>resolver&#45;&gt;live</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M700.89,-94.88C581.65,-88.42 394.43,-78.29 271.7,-71.64"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="271.71,-69.01 264.07,-71.23 271.42,-74.26 271.71,-69.01"/>
<text text-anchor="middle" x="486" y="-92.8" font-family="Arial" font-size="14.00" fill="#536b72">activate</text>
</g>
</g>
</svg>
</div>
```
<!-- /inline-svg -->

::: {.notes}

**Say:** The agent can use packages while worker networking is restricted because preparation is a separate trusted operation. The resolver may download packages and execute install or build code. Its authority is not the worker’s network authority, and automatic resolution must not be described as a sandbox escape.

**Show:** Keep actual YAML and a requirements call above the trust-boundary diagram. Follow the request out to ir/uv and the prepared environment back in; emphasize activation without automatic replay of the cell.

**Sources:** [requirements](https://github.com/t-kalinowski/mcp-console/blob/main/docs/REQUIREMENTS.md) · [architecture](https://github.com/t-kalinowski/mcp-console/blob/main/docs/ARCHITECTURE.md) · [sandbox](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::


## SSH: the same policy, materialized remotely {#ssh .deck-slide .config-slide .code-dense}
::: {.kicker}

Sandbox · configure the boundary

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label .path-label}

.agents/console/config.yaml

:::

```yaml
extends: ":workspace"
target:
  transport:
    kind: ssh
    host: analysis-host
  workspace: /srv/projects/analysis
sandbox:
  filesystem:
    entries:
      - path: {type: path, path: ./data}
        access: read
```

::::

:::::

::::: {.right}

::: {.ownership-card}

Remote workspace and relative paths.

Remote runner and resolver.

Local MCP server and recordings.

:::

:::::

::::::


::: {.notes}

**Say:** The target selects where execution happens; it does not grant permissions. Console reads YAML locally, captures it, and materializes the native policy on the SSH host. Dependencies are prepared there, while the controller retains the session record.

**Show:** Walk through the visible YAML from top to bottom. This is a complete alternative configuration, not a patch to concatenate with the preceding example.

**Author / capture:** analysis-host must be a configured OpenSSH destination; /srv/projects/analysis and data must already exist. SSH+Docker composition is not supported in this revision.

**Sources:** [ssh](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SSH.md) · [sandbox](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::


## Docker: explicit image, workspace, and binds {#docker .deck-slide .config-slide .code-dense}
::: {.kicker}

Sandbox · configure the boundary

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label .path-label}

.agents/console/config.yaml

:::

```yaml
target:
  workspace: /workspace
  compute:
    kind: docker
    image: my-console:analysis
    pull: if_missing
    mounts:
      - source: .
        target: /workspace
        access: read_write
sandbox:
  filesystem: {kind: external-sandbox}
  network: enabled
```

::::

:::::

::::: {.right}

::: {.ownership-card}

Image supplies runtimes and packages.

Project bind is explicitly writable.

Outer Docker boundary owns enforcement.

:::

:::::

::::::
::: {.takeaway}

A writable bind may expose controller records to worker writes.

:::

::: {.notes}

**Say:** An ordinary Docker image is a prepared execution environment. Dynamic preparation is disabled. This example deliberately delegates to the container boundary; it does not retain native metadata protection inside a writable project bind.

**Show:** Walk through the visible YAML from top to bottom. This is a complete alternative configuration, not a patch to concatenate with the preceding example.

**Author / capture:** Build my-console:analysis first. Source paths are interpreted by the selected Docker daemon. Native sandboxing inside Docker is a separate supported selection where host capabilities permit it; no privileged-mode fallback is implied.

**Sources:** [docker](https://github.com/t-kalinowski/mcp-console/blob/main/docs/DOCKER.md) · [sandbox](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::


## Docker Sandbox uses the compute provider {#docker-sandbox .deck-slide .config-slide .code-dense}
::: {.kicker}

Sandbox · configure the boundary

:::

:::::: {.two-col}

::::: {.left}

:::: {.code-panel}

::: {.panel-label .path-label}

.agents/console/config.yaml

:::

```yaml
target:
  workspace: /srv/analysis
  compute:
    kind: docker_sandbox
    template: >-
      registry.example.com/console@sha256:<64-hex-digest>
    mounts:
      - source: /srv/analysis
        target: /srv/analysis
        access: read_write
sandbox:
  provider: compute
  environment:
    ANALYSIS_MODE: interactive
```

::::

:::::

::::: {.right}

::: {.ownership-card}

sbx creates an owned microVM.

Digest-qualified prepared template.

Provider policy, not native-runner YAML.

:::

:::::

::::::


::: {.notes}

**Say:** Docker Sandbox is not ordinary Docker. The compute provider owns its boundary and accepts only environment and inheritance controls under sandbox. The native runner is not invoked on this path, and native filesystem, proxy, or extends settings are rejected.

**Show:** Walk through the visible YAML from top to bottom. This is a complete alternative configuration, not a patch to concatenate with the preceding example.

**Author / capture:** Replace <64-hex-digest> and paths with actual values. Prepare/import the template and provider policy separately. The tested SBX version requires the first shared path to be read/write. No dynamic dependency preparation in the template-backed worker.

**Sources:** [sbx](https://github.com/t-kalinowski/mcp-console/blob/main/docs/DOCKER_SANDBOX.md) · [sandbox](https://github.com/t-kalinowski/mcp-console/blob/main/docs/SANDBOX_CONFIGURATION.md)

:::


## The call stays the same {#target-invariance .deck-slide}
::: {.kicker}

Execution targets

:::

::: {.code-panel}

::: {.panel-label}

send

:::

```python
console.send(r="coef(fit)")
```

:::

::: {.target-line}

Local host &nbsp; / &nbsp; SSH &nbsp; / &nbsp; Docker &nbsp; / &nbsp; Docker Sandbox

:::



::: {.notes}

**Say:** Across targets, the model still submits cells, receives bounded results, and uses the same controls. Available environments and preparation capabilities can differ and the live schema reflects that. The claim is a stable interaction model, not identical permissions or package-management behavior everywhere.

**Show:** Return to one familiar call with four destinations beneath it. This is the bridge into client integrations: both the client above and execution target below can vary without changing the central idea.

**Sources:** [Implemented architecture](https://github.com/t-kalinowski/mcp-console/blob/main/docs/ARCHITECTURE.md) · [Canonical tool schema snapshot](https://github.com/t-kalinowski/mcp-console/blob/main/tests/snapshots/client_server/server/test_tools/initializes_and_lists_tools.yaml) · [Python clients and integrations](https://github.com/t-kalinowski/mcp-console/blob/main/docs/PYTHON.md)

:::



## The process topology {#process-topology .deck-slide}
::: {.kicker}

Architecture

:::

<!-- inline-svg:assets/architecture.svg -->
```{=html}
<div class="embedded-diagram">
<svg role="img" aria-label="MCP client to Rust server, external native sandbox runner, relay, and worker; dependency preparation remains outside the worker boundary." width="1359pt" height="257pt"
 viewBox="0.00 0.00 1359.00 257.00" xmlns="http://www.w3.org/2000/svg" xmlns:xlink="http://www.w3.org/1999/xlink">
<g id="diagram-8-graph0" class="graph" transform="scale(1 1) rotate(0) translate(18 239)">
<title>G</title>
<g id="diagram-8-clust1" class="cluster">
<title>cluster_sandbox</title>
<path fill="transparent" stroke="#167d76" d="M838,-96C838,-96 1303,-96 1303,-96 1309,-96 1315,-102 1315,-108 1315,-108 1315,-201 1315,-201 1315,-207 1309,-213 1303,-213 1303,-213 838,-213 838,-213 832,-213 826,-207 826,-201 826,-201 826,-108 826,-108 826,-102 832,-96 838,-96"/>
<text text-anchor="middle" x="1070.5" y="-193.8" font-family="Arial" font-size="19.00" fill="#263d46">SANDBOXED PROCESSES</text>
</g>
<!-- client -->
<g id="diagram-8-node1" class="node">
<title>client</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M122,-87C122,-87 12,-87 12,-87 6,-87 0,-81 0,-75 0,-75 0,-51 0,-51 0,-45 6,-39 12,-39 12,-39 122,-39 122,-39 128,-39 134,-45 134,-51 134,-51 134,-75 134,-75 134,-81 128,-87 122,-87"/>
<text text-anchor="middle" x="67" y="-57.4" font-family="Arial" font-size="22.00" fill="#142d39">MCP client</text>
</g>
<!-- server -->
<g id="diagram-8-node2" class="node">
<title>server</title>
<path fill="#d9eee9" stroke="#167d76" stroke-width="1.5" d="M438,-99C438,-99 260,-99 260,-99 254,-99 248,-93 248,-87 248,-87 248,-39 248,-39 248,-33 254,-27 260,-27 260,-27 438,-27 438,-27 444,-27 450,-33 450,-39 450,-39 450,-87 450,-87 450,-93 444,-99 438,-99"/>
<text text-anchor="middle" x="349" y="-69.4" font-family="Arial" font-size="22.00" fill="#142d39">Rust server</text>
<text text-anchor="middle" x="349" y="-45.4" font-family="Arial" font-size="22.00" fill="#142d39">session + records</text>
</g>
<!-- client&#45;&gt;server -->
<g id="diagram-8-edge2" class="edge">
<title>client&#45;&gt;server</title>
<path fill="none" stroke="#4c7571" stroke-width="1.7" d="M134,-63C165.83,-63 204.82,-63 240.77,-63"/>
<polygon fill="#4c7571" stroke="#4c7571" stroke-width="1.7" points="240.91,-65.45 247.91,-63 240.91,-60.55 240.91,-65.45"/>
<text text-anchor="middle" x="191" y="-68" font-family="Arial" font-size="15.00" fill="#445b64">MCP stdio</text>
</g>
<!-- runner -->
<g id="diagram-8-node3" class="node">
<title>runner</title>
<path fill="#f5e7dc" stroke="#b56844" stroke-width="1.5" d="M777,-176C777,-176 558,-176 558,-176 552,-176 546,-170 546,-164 546,-164 546,-116 546,-116 546,-110 552,-104 558,-104 558,-104 777,-104 777,-104 783,-104 789,-110 789,-116 789,-116 789,-164 789,-164 789,-170 783,-176 777,-176"/>
<text text-anchor="middle" x="667.5" y="-146.4" font-family="Arial" font-size="22.00" fill="#142d39">External native runner</text>
<text text-anchor="middle" x="667.5" y="-122.4" font-family="Arial" font-size="22.00" fill="#142d39">policy + supervision</text>
</g>
<!-- server&#45;&gt;runner -->
<g id="diagram-8-edge3" class="edge">
<title>server&#45;&gt;runner</title>
<path fill="none" stroke="#4c7571" stroke-width="1.7" d="M450.14,-87.36C478.31,-94.22 509.33,-101.76 538.87,-108.95"/>
<polygon fill="#4c7571" stroke="#4c7571" stroke-width="1.7" points="538.56,-111.4 545.94,-110.67 539.72,-106.63 538.56,-111.4"/>
<text text-anchor="middle" x="498" y="-108" font-family="Arial" font-size="15.00" fill="#445b64">launch</text>
</g>
<!-- resolver -->
<g id="diagram-8-node6" class="node">
<title>resolver</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M768.5,-72C768.5,-72 566.5,-72 566.5,-72 560.5,-72 554.5,-66 554.5,-60 554.5,-60 554.5,-12 554.5,-12 554.5,-6 560.5,0 566.5,0 566.5,0 768.5,0 768.5,0 774.5,0 780.5,-6 780.5,-12 780.5,-12 780.5,-60 780.5,-60 780.5,-66 774.5,-72 768.5,-72"/>
<text text-anchor="middle" x="667.5" y="-42.4" font-family="Arial" font-size="22.00" fill="#142d39">ir / uv resolvers</text>
<text text-anchor="middle" x="667.5" y="-18.4" font-family="Arial" font-size="22.00" fill="#142d39">trusted host process</text>
</g>
<!-- server&#45;&gt;resolver -->
<g id="diagram-8-edge5" class="edge">
<title>server&#45;&gt;resolver</title>
<path fill="none" stroke="#4c7571" stroke-width="1.7" stroke-dasharray="5,2" d="M450.14,-54.46C480.95,-51.83 515.18,-48.91 547.15,-46.18"/>
<polygon fill="#4c7571" stroke="#4c7571" stroke-width="1.7" points="547.59,-48.6 554.36,-45.57 547.17,-43.72 547.59,-48.6"/>
<text text-anchor="middle" x="498" y="-56" font-family="Arial" font-size="15.00" fill="#445b64">prepare</text>
</g>
<!-- relay -->
<g id="diagram-8-node4" class="node">
<title>relay</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M994,-176C994,-176 846,-176 846,-176 840,-176 834,-170 834,-164 834,-164 834,-116 834,-116 834,-110 840,-104 846,-104 846,-104 994,-104 994,-104 1000,-104 1006,-110 1006,-116 1006,-116 1006,-164 1006,-164 1006,-170 1000,-176 994,-176"/>
<text text-anchor="middle" x="920" y="-146.4" font-family="Arial" font-size="22.00" fill="#142d39">Relay</text>
<text text-anchor="middle" x="920" y="-122.4" font-family="Arial" font-size="22.00" fill="#142d39">code / streams</text>
</g>
<!-- runner&#45;&gt;relay -->
<g id="diagram-8-edge4" class="edge">
<title>runner&#45;&gt;relay</title>
<path fill="none" stroke="#4c7571" stroke-width="1.7" d="M789.35,-140C801.9,-140 814.51,-140 826.67,-140"/>
<polygon fill="#4c7571" stroke="#4c7571" stroke-width="1.7" points="826.78,-142.45 833.78,-140 826.78,-137.55 826.78,-142.45"/>
</g>
<!-- worker -->
<g id="diagram-8-node5" class="node">
<title>worker</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M1295,-176C1295,-176 1123,-176 1123,-176 1117,-176 1111,-170 1111,-164 1111,-164 1111,-116 1111,-116 1111,-110 1117,-104 1123,-104 1123,-104 1295,-104 1295,-104 1301,-104 1307,-110 1307,-116 1307,-116 1307,-164 1307,-164 1307,-170 1301,-176 1295,-176"/>
<text text-anchor="middle" x="1209" y="-146.4" font-family="Arial" font-size="22.00" fill="#142d39">Worker</text>
<text text-anchor="middle" x="1209" y="-122.4" font-family="Arial" font-size="22.00" fill="#142d39">R · Python · SQL</text>
</g>
<!-- relay&#45;&gt;worker -->
<g id="diagram-8-edge1" class="edge">
<title>relay&#45;&gt;worker</title>
<path fill="none" stroke="#4c7571" stroke-width="1.7" d="M1006.33,-140C1036.9,-140 1071.71,-140 1103.7,-140"/>
<polygon fill="#4c7571" stroke="#4c7571" stroke-width="1.7" points="1103.91,-142.45 1110.91,-140 1103.91,-137.55 1103.91,-142.45"/>
<text text-anchor="middle" x="1058.5" y="-162" font-family="Arial" font-size="15.00" fill="#445b64">sideband</text>
<text text-anchor="middle" x="1058.5" y="-145" font-family="Arial" font-size="15.00" fill="#445b64">+ stdio</text>
</g>
</g>
</svg>
</div>
```
<!-- /inline-svg -->

::: {.takeaway}

Client → server → external runner → relay → worker.

:::

::: {.notes}

**Say:** Show the whole architecture after its consequences are familiar. The MCP client talks to the Rust server. The execution launcher establishes the selected target and boundary; the relay owns worker communication and direct supervision; the worker owns language state. Dependency preparation is outside the worker sandbox. Native helper details differ by platform.

**Show:** This is the main architecture diagram. Keep the process boundary visible and distinguish the client, controller, and worker. The launcher node is an abstraction: native execution uses the sandbox runner, while compute targets have their own owners and provider behavior.

**Sources:** [Implemented architecture](https://github.com/t-kalinowski/mcp-console/blob/main/docs/ARCHITECTURE.md)

:::



## State has an owner {#state-ownership .deck-slide}
::: {.kicker}

Architecture

:::

::::: {.two-col}

:::: {.left}

::: {.ownership-card}

<h3>Server</h3>

Session contract<br>
Retained requirements<br>
Output and recordings<br>
Replacement coordination

:::

::::

:::: {.right}

::: {.ownership-card}

<h3>Worker</h3>

R and Python objects<br>
Database state<br>
Debugger and input state<br>
Language evaluation

:::

::::

:::::



::: {.notes}

**Say:** The architecture becomes easier to reason about when state ownership is explicit. The server owns logical-session and delivery state; the worker owns live computational state. That separation explains why prepared requirements and recordings survive worker restart while objects and debugger state do not.

**Show:** Use a two-column owner table without method names or module-level details. Tie it directly back to the restart behavior already demonstrated. Do not introduce a new peer-runtime coordinator as implemented unless that refactor has landed.

**Sources:** [Implemented architecture](https://github.com/t-kalinowski/mcp-console/blob/main/docs/ARCHITECTURE.md) · [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md)

:::



## Three worker paths, not prompt scraping {#worker-streams .deck-slide}
::: {.kicker}

Architecture

:::

<!-- inline-svg:assets/streams.svg -->
```{=html}
<div class="embedded-diagram">
<svg role="img" aria-label="The relay uses a sideband for cells and control, fd 0 for interactive input, and fd 1/2 for raw output." width="643pt" height="105pt"
 viewBox="0.00 0.00 642.80 105.33" xmlns="http://www.w3.org/2000/svg" xmlns:xlink="http://www.w3.org/1999/xlink">
<g id="diagram-9-graph0" class="graph" transform="scale(1 1) rotate(0) translate(14.4 90.93)">
<title>G</title>
<!-- server -->
<g id="diagram-9-node1" class="node">
<title>server</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M73,-63.53C73,-63.53 12,-63.53 12,-63.53 6,-63.53 0,-57.53 0,-51.53 0,-51.53 0,-29.53 0,-29.53 0,-23.53 6,-17.53 12,-17.53 12,-17.53 73,-17.53 73,-17.53 79,-17.53 85,-23.53 85,-29.53 85,-29.53 85,-51.53 85,-51.53 85,-57.53 79,-63.53 73,-63.53"/>
<text text-anchor="middle" x="42.5" y="-36.13" font-family="Arial" font-size="18.00" fill="#142d39">Server</text>
</g>
<!-- relay -->
<g id="diagram-9-node2" class="node">
<title>relay</title>
<path fill="#d9eee9" stroke="#167d76" stroke-width="1.5" d="M318,-63.53C318,-63.53 264,-63.53 264,-63.53 258,-63.53 252,-57.53 252,-51.53 252,-51.53 252,-29.53 252,-29.53 252,-23.53 258,-17.53 264,-17.53 264,-17.53 318,-17.53 318,-17.53 324,-17.53 330,-23.53 330,-29.53 330,-29.53 330,-51.53 330,-51.53 330,-57.53 324,-63.53 318,-63.53"/>
<text text-anchor="middle" x="291" y="-36.13" font-family="Arial" font-size="18.00" fill="#142d39">Relay</text>
</g>
<!-- server&#45;&gt;relay -->
<g id="diagram-9-edge1" class="edge">
<title>server&#45;&gt;relay</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M92.65,-40.53C136.87,-40.53 201.33,-40.53 244.42,-40.53"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="92.54,-37.9 85.04,-40.53 92.54,-43.15 92.54,-37.9"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="244.44,-43.15 251.94,-40.53 244.44,-37.9 244.44,-43.15"/>
<text text-anchor="middle" x="168.5" y="-59.33" font-family="Arial" font-size="14.00" fill="#536b72">commands / events</text>
<text text-anchor="middle" x="168.5" y="-44.33" font-family="Arial" font-size="14.00" fill="#536b72">stdio JSONL</text>
</g>
<!-- worker -->
<g id="diagram-9-node3" class="node">
<title>worker</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M602,-63.53C602,-63.53 536,-63.53 536,-63.53 530,-63.53 524,-57.53 524,-51.53 524,-51.53 524,-29.53 524,-29.53 524,-23.53 530,-17.53 536,-17.53 536,-17.53 602,-17.53 602,-17.53 608,-17.53 614,-23.53 614,-29.53 614,-29.53 614,-51.53 614,-51.53 614,-57.53 608,-63.53 602,-63.53"/>
<text text-anchor="middle" x="569" y="-36.13" font-family="Arial" font-size="18.00" fill="#142d39">Worker</text>
</g>
<!-- relay&#45;&gt;worker -->
<g id="diagram-9-edge2" class="edge">
<title>relay&#45;&gt;worker</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M337.52,-53.66C343.03,-54.84 348.62,-55.84 354,-56.53 418.37,-64.7 435.54,-64.01 500,-56.53 505.3,-55.91 510.79,-55.03 516.24,-54"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="337.97,-51.07 330.07,-51.96 336.8,-56.19 337.97,-51.07"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="516.8,-56.57 523.64,-52.52 515.77,-51.42 516.8,-56.57"/>
<text text-anchor="middle" x="427" y="-65.33" font-family="Arial" font-size="14.00" fill="#536b72">cells + control: sideband</text>
</g>
<!-- relay&#45;&gt;worker -->
<g id="diagram-9-edge3" class="edge">
<title>relay&#45;&gt;worker</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M330.01,-36.97C337.93,-36.37 346.23,-35.84 354,-35.53 418.84,-32.92 435.15,-33.14 500,-35.53 505.17,-35.71 510.55,-35.99 515.92,-36.3"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="516.12,-38.95 523.77,-36.8 516.45,-33.71 516.12,-38.95"/>
<text text-anchor="middle" x="427" y="-39.33" font-family="Arial" font-size="14.00" fill="#536b72">interactive input: fd 0</text>
</g>
<!-- worker&#45;&gt;relay -->
<g id="diagram-9-edge4" class="edge">
<title>worker&#45;&gt;relay</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M523.89,-17.94C516.11,-14.87 507.95,-12.2 500,-10.53 436.51,2.86 417.25,3.97 354,-10.53 348.35,-11.82 342.61,-13.7 337.02,-15.9"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="335.94,-13.51 330.04,-18.84 337.97,-18.35 335.94,-13.51"/>
<text text-anchor="middle" x="427" y="-14.33" font-family="Arial" font-size="14.00" fill="#536b72">raw output: fd 1 + fd 2</text>
</g>
</g>
</svg>
</div>
```
<!-- /inline-svg -->



::: {.notes}

**Say:** Code/control messages, interactive input, and raw output have distinct channels. The relay keeps these concerns separate. Console does not have to guess whether a program is ready by matching a printed greater-than sign. Structured runtime state is part of why timeout, debugger, and failure behavior can have precise contracts.

**Show:** Use the stream diagram. It should make the distinction introduced by stdin visually concrete. This is not a deep dive into JSONL framing or file-descriptor bootstrap details.

**Sources:** [Implemented architecture](https://github.com/t-kalinowski/mcp-console/blob/main/docs/ARCHITECTURE.md) · [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [Vision / intended design](https://github.com/t-kalinowski/mcp-console/blob/main/design-sketches/VISION.md)

:::



## ellmer {#ellmer .deck-slide .integration}
::: {.kicker}

Use your existing client

:::

::: {.code-panel}

::: {.panel-label}

R · register one tool

:::

```r
library(ellmer)
library(mcp.console)

chat <- chat_openai()
chat$register_tool(console_tool())
chat$chat("Analyze measurements.csv with Console.")
```

:::

::: {.takeaway}

Your chat and provider. A separate persistent Console worker.

:::

::: {.notes}

**Say:** Give ellmer one slide. The R wrapper starts Console, reads its live tool schema, and presents it as an ellmer tool. It does not execute agent code in the R process running the chat. Provider choice remains with ellmer and the application. Keep the tool alive across the conversation so its session persists.

**Show:** Show only the registration example, with the registration line emphasized. No installation explanation here; that is the next section. The author notes include the asynchronous sequential-tool caveat from the R README.

**Sources:** [R package interface](https://github.com/t-kalinowski/mcp-console/blob/main/r/README.md) · [R tool wrapper implementation](https://github.com/t-kalinowski/mcp-console/blob/main/r/R/console-tool.R)

:::



## chatlas {#chatlas .deck-slide .integration}
::: {.kicker}

Use your existing client

:::

::: {.code-panel}

::: {.panel-label}

Python · preserve the server-provided schema

:::

```{.python code-line-numbers="2,5-7"}
from chatlas import ChatOpenAI
import mcp_console

chat = ChatOpenAI()
with mcp_console.MCPConsole() as console:
    tool = mcp_console.chatlas.tool(console)
    chat.set_tools([*chat.get_tools(), tool])
    chat.chat("Analyze measurements.csv with Console.")
```

:::

::: {.takeaway}

The same registration pattern in Python.

:::

::: {.notes}

**Say:** The chatlas adapter adds the same tool to an existing chat. Use the adapter and set_tools so the explicit server schema is preserved rather than inferred from a generic Python function annotation. The application still owns the chat loop and the Console connection lifetime.

**Show:** Keep the visual design identical to the ellmer slide. For a visual demo, verify image preservation on the chosen adapter path; the native MCP registration variant is included in the examples file.

**Sources:** [Python clients and integrations](https://github.com/t-kalinowski/mcp-console/blob/main/docs/PYTHON.md)

:::



## Register Console with your CLI client {#codex .deck-slide .integration}
::: {.kicker}

Use your existing client

:::

:::: {.code-panel}

::: {.panel-label}

Codex CLI

:::

```bash
codex mcp add console -- uvx mcp-console serve
```

::::
:::: {.code-panel}

::: {.panel-label}

Claude Code CLI · the same Console installation

:::

```bash
claude mcp add console -- uvx mcp-console serve
```

::::

::: {.notes}

**Say:** Both clients can launch the same Console server over MCP stdio. Run the registration command for the client you use. Client-side deadlines remain separate from Console’s wait timeout.

**Show:** The launcher and server arguments are identical. The commands were checked against the installed CLI help; no configuration was changed.

**Sources:** [Codex MCP](https://developers.openai.com/codex/mcp/) · [Claude Code MCP](https://code.claude.com/docs/en/mcp)

:::

## Anthropic Python SDK {#anthropic .deck-slide .integration .code-dense}
::: {.kicker}

Use your existing client

:::

::: {.code-panel}

::: {.panel-label}

Native MCP tools · inside your async application

:::

```{.python code-line-numbers="2,5"}
from anthropic import AsyncAnthropic
import mcp_console

client = AsyncAnthropic()
async with mcp_console.anthropic.tools() as tools:
    runner = client.beta.messages.tool_runner(
        model=model_id,
        max_tokens=4096,
        tools=tools,
        messages=[{"role": "user", "content": prompt}],
    )
    async for message in runner:
        print(message)
```

:::

::: {.takeaway}

Use the native MCP tool path when the workflow includes images.

:::

::: {.notes}

**Say:** The Anthropic integration has native MCP and function-tool paths. Show the native async tools context here because visual workflows need image preservation. The SDK owns the model loop; Console owns execution. model_id and prompt are application inputs, not hard-coded recommendations.

**Show:** This is an async application excerpt, with the full runnable wrapper in examples/anthropic_client.py. Keep the connection open for the run and let the context close it. The concrete API follows the repository’s documented integration and should be rechecked before presentation.

**Sources:** [Python clients and integrations](https://github.com/t-kalinowski/mcp-console/blob/main/docs/PYTHON.md)

:::



## The direct Python client {#python-client .deck-slide .integration}
::: {.kicker}

Use your existing client

:::

::: {.code-panel}

::: {.panel-label}

Synchronous client

:::

```{.python code-line-numbers="1,3-4,7"}
import mcp_console

with mcp_console.MCPConsole() as console:
    result = console.send(r="sum(1:10)")
    print(result)

    result = console.send(python="sum(range(11))")
    print(result)
```

:::



::: {.notes}

**Say:** The direct client is useful for scripts, application integration, and deterministic checks. Both calls reuse one Console connection. The convenience return is text and uses placeholders for image content; native MCP or image-preserving adapters are the appropriate interface for a visual agent workflow. AsyncMCPConsole offers the parallel asynchronous API.

**Show:** Show a complete minimal synchronous example. Do not imply that printing the direct client result displays a plot. Keep the async counterpart in the accompanying example file and the compatibility list.

**Sources:** [Python clients and integrations](https://github.com/t-kalinowski/mcp-console/blob/main/docs/PYTHON.md)

:::



## OpenAI Responses {#openai-responses .deck-slide .integration .code-dense}
::: {.kicker}

Use your existing client

:::

::: {.code-panel}

::: {.panel-label}

Registration excerpt · the application owns the tool loop

:::

```{.python code-line-numbers="2,5-6"}
from openai import OpenAI
import mcp_console

client = OpenAI()
with mcp_console.MCPConsole() as console:
    tool = mcp_console.openai.responses_tool(console)
    response = client.responses.create(
        model=model_id,
        input=prompt,
        tools=[tool.definition],
    )
    # Route function calls through tool(call).
    # Full continuation loop: examples/openai_responses.py
```

:::

::: {.takeaway}

Live server schema in; text and image results out.

:::

::: {.notes}

**Say:** The Responses adapter reads the connected server schema and supplies the function definition and function-call output conversion. The application still owns the continuation loop. This excerpt deliberately focuses on registration; the complete loop is provided as a companion example rather than squeezed onto the slide.

**Show:** Show the adapter line as the focus. model_id and prompt are supplied by the host application. The full example includes continuation and a turn limit; do not present this excerpt alone as a complete autonomous agent run.

**Sources:** [Python clients and integrations](https://github.com/t-kalinowski/mcp-console/blob/main/docs/PYTHON.md)

:::



## Other supported entry points {#other-clients .deck-slide .compatibility}
::: {.kicker}

Use your existing client

:::

::: {.compat-list}

- `AsyncMCPConsole`
- OpenAI Agents SDK
- Codex Python thread SDK
- Native MCP registration in chatlas
- Compatible MCP stdio clients

:::

::: {.takeaway}

Use the live schema; keep the session connection alive.

:::

::: {.small-note}

Image handling and client deadlines are integration-specific.

:::

::: {.notes}

**Say:** Finish the client sequence with one list, not another onboarding section. These are the additional entry points documented by the project. Avoid the unqualified phrase works with everything: compatibility depends on the client’s transport, tool-schema handling, image support, and timeout behavior.

**Show:** Use a plain list of exact interface names. The source package contains a concise example for each documented integration family; there is no need to display every tool loop in the main talk.

**Sources:** [Python clients and integrations](https://github.com/t-kalinowski/mcp-console/blob/main/docs/PYTHON.md) · [Canonical tool schema snapshot](https://github.com/t-kalinowski/mcp-console/blob/main/tests/snapshots/client_server/server/test_tools/initializes_and_lists_tools.yaml)

:::



## Now, installation from R {#install-r .deck-slide}
::: {.kicker}

Installation

:::

::: {.code-panel}

::: {.panel-label}

R package entry point

:::

```r
install.packages("mcp.console")
```

:::

::: {.takeaway}

Install the R interface. The application is obtained automatically when needed.

:::

::: {.notes}

**Say:** Only after the value and client examples are clear, show the R installation story. For presentation day, assume mcp.console is published to the intended R package repository. Installing the R package gives the interface; creating the tool resolves the Console executable when necessary. Do not say the PyPI download occurs during install.packages itself.

**Show:** Show one large line. This is explicitly an intended publication pathway: the currently inspected R README uses a GitHub installation command. Keep the present-day command in the author checklist, not as competing text on this slide.

**Author check before presenting:** Target-day assumption: mcp.console is published to an R repository usable by install.packages(). Current README documents pak::pak("github::t-kalinowski/mcp-console/r").

**Sources:** [R package interface](https://github.com/t-kalinowski/mcp-console/blob/main/r/README.md)

:::



## The R wrapper obtains the native application {#install-chain .deck-slide}
::: {.kicker}

Installation

:::

<!-- inline-svg:assets/install.svg -->
```{=html}
<div class="embedded-diagram">
<svg role="img" aria-label="The R wrapper uses reticulate uv_run_tool on first use to obtain a cached PyPI platform wheel and native bundle." width="1125pt" height="105pt"
 viewBox="0.00 0.00 1124.80 104.80" xmlns="http://www.w3.org/2000/svg" xmlns:xlink="http://www.w3.org/1999/xlink">
<g id="diagram-10-graph0" class="graph" transform="scale(1 1) rotate(0) translate(14.4 90.4)">
<title>G</title>
<!-- r -->
<g id="diagram-10-node1" class="node">
<title>r</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M150,-76C150,-76 12,-76 12,-76 6,-76 0,-70 0,-64 0,-64 0,-12 0,-12 0,-6 6,0 12,0 12,0 150,0 150,0 156,0 162,-6 162,-12 162,-12 162,-64 162,-64 162,-70 156,-76 150,-76"/>
<text text-anchor="middle" x="81" y="-44.6" font-family="Arial" font-size="23.00" fill="#142d39">mcp.console</text>
<text text-anchor="middle" x="81" y="-19.6" font-family="Arial" font-size="23.00" fill="#142d39">R wrapper</text>
</g>
<!-- uv -->
<g id="diagram-10-node2" class="node">
<title>uv</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M414,-76C414,-76 270,-76 270,-76 264,-76 258,-70 258,-64 258,-64 258,-12 258,-12 258,-6 264,0 270,0 270,0 414,0 414,0 420,0 426,-6 426,-12 426,-12 426,-64 426,-64 426,-70 420,-76 414,-76"/>
<text text-anchor="middle" x="342" y="-44.6" font-family="Arial" font-size="23.00" fill="#142d39">uv_run_tool()</text>
<text text-anchor="middle" x="342" y="-19.6" font-family="Arial" font-size="23.00" fill="#142d39">via reticulate</text>
</g>
<!-- r&#45;&gt;uv -->
<g id="diagram-10-edge1" class="edge">
<title>r&#45;&gt;uv</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M162.18,-38C190,-38 221.39,-38 250.02,-38"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="250.12,-40.63 257.62,-38 250.12,-35.38 250.12,-40.63"/>
<text text-anchor="middle" x="210" y="-41.8" font-family="Arial" font-size="14.00" fill="#536b72">first use</text>
</g>
<!-- pypi -->
<g id="diagram-10-node3" class="node">
<title>pypi</title>
<path fill="#d9eee9" stroke="#167d76" stroke-width="1.5" d="M741,-76C741,-76 582,-76 582,-76 576,-76 570,-70 570,-64 570,-64 570,-12 570,-12 570,-6 576,0 582,0 582,0 741,0 741,0 747,0 753,-6 753,-12 753,-12 753,-64 753,-64 753,-70 747,-76 741,-76"/>
<text text-anchor="middle" x="661.5" y="-44.6" font-family="Arial" font-size="23.00" fill="#142d39">PyPI</text>
<text text-anchor="middle" x="661.5" y="-19.6" font-family="Arial" font-size="23.00" fill="#142d39">platform wheel</text>
</g>
<!-- uv&#45;&gt;pypi -->
<g id="diagram-10-edge2" class="edge">
<title>uv&#45;&gt;pypi</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M426,-38C467.78,-38 518.69,-38 562.38,-38"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="562.39,-40.63 569.89,-38 562.39,-35.38 562.39,-40.63"/>
<text text-anchor="middle" x="498" y="-41.8" font-family="Arial" font-size="14.00" fill="#536b72">resolve + cache</text>
</g>
<!-- bundle -->
<g id="diagram-10-node4" class="node">
<title>bundle</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M1084,-76C1084,-76 814,-76 814,-76 808,-76 802,-70 802,-64 802,-64 802,-12 802,-12 802,-6 808,0 814,0 814,0 1084,0 1084,0 1090,0 1096,-6 1096,-12 1096,-12 1096,-64 1096,-64 1096,-70 1090,-76 1084,-76"/>
<text text-anchor="middle" x="949" y="-44.6" font-family="Arial" font-size="23.00" fill="#142d39">Native Console</text>
<text text-anchor="middle" x="949" y="-19.6" font-family="Arial" font-size="23.00" fill="#142d39">application + companions</text>
</g>
<!-- pypi&#45;&gt;bundle -->
<g id="diagram-10-edge3" class="edge">
<title>pypi&#45;&gt;bundle</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M753.23,-38C766.38,-38 780.22,-38 794.2,-38"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="794.48,-40.63 801.98,-38 794.48,-35.38 794.48,-40.63"/>
</g>
</g>
</svg>
</div>
```
<!-- /inline-svg -->

::: {.small-note}

Explicit `path=` or `version=` takes precedence; an existing PATH executable can be reused.

:::

::: {.notes}

**Say:** With no explicit selection, console_tool() first uses an mcp-console executable on PATH. If none is found, it resolves the published application with reticulate::uv_run_tool(). A named path or version gives explicit control. PyPI is the distribution channel for a native Rust application bundle, not a requirement for the R user to manually manage an analysis venv.

**Show:** Use the first-use chain. The installation trigger belongs on the arrow from wrapper to resolver. Keep executable selection separate from analysis runtime environment selection, which was covered earlier.

**Sources:** [R package interface](https://github.com/t-kalinowski/mcp-console/blob/main/r/README.md) · [R tool wrapper implementation](https://github.com/t-kalinowski/mcp-console/blob/main/r/R/console-tool.R) · [Release and installed bundle](https://github.com/t-kalinowski/mcp-console/blob/main/RELEASE.md)

:::



## Python or a standalone MCP launch {#install-python .deck-slide .integration}
::: {.kicker}

Installation

:::

::: {.code-panel}

::: {.panel-label}

Python client

:::

```bash
pip install "mcp-console[client]"
```

:::

::: {.code-panel}

::: {.panel-label}

One-shot tool launch

:::

```bash
uvx mcp-console serve
```

:::

::: {.code-panel}

::: {.panel-label}

Persistent command installation

:::

```bash
uv tool install mcp-console
```

:::

::: {.notes}

**Say:** Python users choose the base package or the extra for their integration. MCP clients can launch through uvx, or use a persistent uv tool installation. These are alternative routes into the same distribution, not three installation steps. Framework extras avoid installing every SDK when only the executable is needed.

**Show:** Show three short alternatives with explicit labels. Mention prerequisites only as needed: Python packaging and supported native wheels are not the same thing as the language runtime installed on a selected execution host.

**Sources:** [Project README](https://github.com/t-kalinowski/mcp-console/blob/main/README.md) · [Python clients and integrations](https://github.com/t-kalinowski/mcp-console/blob/main/docs/PYTHON.md)

:::



## One distribution; separate execution environments {#native-bundle .deck-slide}
::: {.kicker}

Installation

:::

::::: {.two-col}

:::: {.left}

::: {.big-statement}

PyPI wheel<br>Native application + companions

:::

::::

:::: {.right}

::: {.big-statement}

Execution host<br>Selected runtimes + packages

:::

::::

:::::

::: {.takeaway}

Supported wheels ship native binaries; execution hosts supply the runtimes.

:::

::: {.notes}

**Say:** A supported platform wheel carries the application and sandbox companions as an installed bundle. Language runtimes and analysis packages belong to the execution environment. Current support is macOS and Linux, with specific Linux glibc and sandbox prerequisites; Windows is not currently supported. Source builds still need the toolchain. An SSH or compute controller does not require its own local analysis environment.

**Show:** Use the two responsibility boxes. Do not put a green all-platforms checkmark on the slide. Keep exact release/platform requirements in the installation documentation, because those can change before presentation day.

**Sources:** [Project README](https://github.com/t-kalinowski/mcp-console/blob/main/README.md) · [Release and installed bundle](https://github.com/t-kalinowski/mcp-console/blob/main/RELEASE.md) · [Implemented architecture](https://github.com/t-kalinowski/mcp-console/blob/main/docs/ARCHITECTURE.md)

:::



## Interactive work, without a sprawling tool surface {#closing .deck-slide .hero}
::: {.kicker}

Closing

:::

::: {.big-statement}

One `console.send()`.<br>A full interactive runtime.

:::

::: {.takeaway}

Small routine calls. Explicit controls. Reusable state. A record of the work.

:::

::: {.language-line}

R · Python · SQL &nbsp; / &nbsp; Local · Remote · Container · MicroVM

:::

::: {.notes}

**Say:** Close on the design, not on installation commands. The contribution is a compact model-facing interface with enough runtime capability and explicit control to remain useful as the model’s ability grows. The interaction carries across languages, clients, and execution targets while keeping ownership and trust boundaries visible.

**Show:** Return to the opening typography, now with the expanded capability set condensed into one line. Stop the main presentation here; the development-practices slides follow as an addendum.

**Sources:** [Canonical tool schema snapshot](https://github.com/t-kalinowski/mcp-console/blob/main/tests/snapshots/client_server/server/test_tools/initializes_and_lists_tools.yaml) · [Built-in runtime](https://github.com/t-kalinowski/mcp-console/blob/main/docs/BUILTIN_RUNTIME.md) · [Implemented architecture](https://github.com/t-kalinowski/mcp-console/blob/main/docs/ARCHITECTURE.md)

:::



## Rust implementation. Python integration tests. {#tests-python .deck-slide}
::: {.kicker}

Addendum · Development practices

:::

::::: {.two-col}

:::: {.left}

::: {.big-statement}

Implement in Rust

:::

::::

:::: {.right}

::: {.big-statement}

Exercise from Python

:::

::::

:::::



::: {.notes}

**Say:** The main integration suite is Python even though the application is Rust. It drives the built executable and observes the real process boundary. Small unit tests can still own pure parsing or validation policy; the claim is not that absolutely no Rust tests exist.

**Show:** Use two simple implementation/test boxes. Lead with why this matters: the test language can stay independent of implementation refactors while validating the product contract users and clients actually see.

**Sources:** [Boundary test guide](https://github.com/t-kalinowski/mcp-console/blob/main/tests/boundaries/README.md) · [Project README](https://github.com/t-kalinowski/mcp-console/blob/main/README.md)

:::



## Test the boundaries you intend to preserve {#test-boundaries .deck-slide}
::: {.kicker}

Addendum · Development practices

:::

<!-- inline-svg:assets/boundaries.svg -->
```{=html}
<div class="embedded-diagram">
<svg role="img" aria-label="Tests observe client-server, server-relay, relay-worker, and CLI boundaries." width="855pt" height="153pt"
 viewBox="0.00 0.00 854.80 152.80" xmlns="http://www.w3.org/2000/svg" xmlns:xlink="http://www.w3.org/1999/xlink">
<g id="diagram-11-graph0" class="graph" transform="scale(1 1) rotate(0) translate(14.4 138.4)">
<title>G</title>
<!-- client -->
<g id="diagram-11-node1" class="node">
<title>client</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M125,-124C125,-124 37,-124 37,-124 31,-124 25,-118 25,-112 25,-112 25,-90 25,-90 25,-84 31,-78 37,-78 37,-78 125,-78 125,-78 131,-78 137,-84 137,-90 137,-90 137,-112 137,-112 137,-118 131,-124 125,-124"/>
<text text-anchor="middle" x="81" y="-96.6" font-family="Arial" font-size="18.00" fill="#142d39">Test client</text>
</g>
<!-- server -->
<g id="diagram-11-node2" class="node">
<title>server</title>
<path fill="#d9eee9" stroke="#167d76" stroke-width="1.5" d="M397,-85C397,-85 299,-85 299,-85 293,-85 287,-79 287,-73 287,-73 287,-51 287,-51 287,-45 293,-39 299,-39 299,-39 397,-39 397,-39 403,-39 409,-45 409,-51 409,-51 409,-73 409,-73 409,-79 403,-85 397,-85"/>
<text text-anchor="middle" x="348" y="-57.6" font-family="Arial" font-size="18.00" fill="#142d39">Built server</text>
</g>
<!-- client&#45;&gt;server -->
<g id="diagram-11-edge1" class="edge">
<title>client&#45;&gt;server</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M137.23,-92.87C178.53,-86.79 235.22,-78.45 279.25,-71.97"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="279.73,-74.55 286.76,-70.86 278.96,-69.36 279.73,-74.55"/>
<text text-anchor="middle" x="224.5" y="-87.8" font-family="Arial" font-size="14.00" fill="#536b72">client_server</text>
</g>
<!-- relay -->
<g id="diagram-11-node3" class="node">
<title>relay</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M598,-85C598,-85 544,-85 544,-85 538,-85 532,-79 532,-73 532,-73 532,-51 532,-51 532,-45 538,-39 544,-39 544,-39 598,-39 598,-39 604,-39 610,-45 610,-51 610,-51 610,-73 610,-73 610,-79 604,-85 598,-85"/>
<text text-anchor="middle" x="571" y="-57.6" font-family="Arial" font-size="18.00" fill="#142d39">Relay</text>
</g>
<!-- server&#45;&gt;relay -->
<g id="diagram-11-edge2" class="edge">
<title>server&#45;&gt;relay</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M409.1,-62C445.31,-62 490.71,-62 524.04,-62"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="524.31,-64.63 531.81,-62 524.31,-59.38 524.31,-64.63"/>
<text text-anchor="middle" x="470.5" y="-65.8" font-family="Arial" font-size="14.00" fill="#536b72">server_relay</text>
</g>
<!-- worker -->
<g id="diagram-11-node4" class="node">
<title>worker</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M814,-85C814,-85 748,-85 748,-85 742,-85 736,-79 736,-73 736,-73 736,-51 736,-51 736,-45 742,-39 748,-39 748,-39 814,-39 814,-39 820,-39 826,-45 826,-51 826,-51 826,-73 826,-73 826,-79 820,-85 814,-85"/>
<text text-anchor="middle" x="781" y="-57.6" font-family="Arial" font-size="18.00" fill="#142d39">Worker</text>
</g>
<!-- relay&#45;&gt;worker -->
<g id="diagram-11-edge3" class="edge">
<title>relay&#45;&gt;worker</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M610.15,-62C643.33,-62 691.69,-62 728.33,-62"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="728.46,-64.63 735.96,-62 728.46,-59.38 728.46,-64.63"/>
<text text-anchor="middle" x="673" y="-65.8" font-family="Arial" font-size="14.00" fill="#536b72">relay_worker</text>
</g>
<!-- cli -->
<g id="diagram-11-node5" class="node">
<title>cli</title>
<path fill="#ffffff" stroke="#a9b6b6" stroke-width="1.5" d="M150,-46C150,-46 12,-46 12,-46 6,-46 0,-40 0,-34 0,-34 0,-12 0,-12 0,-6 6,0 12,0 12,0 150,0 150,0 156,0 162,-6 162,-12 162,-12 162,-34 162,-34 162,-40 156,-46 150,-46"/>
<text text-anchor="middle" x="81" y="-18.6" font-family="Arial" font-size="18.00" fill="#142d39">CLI test process</text>
</g>
<!-- cli&#45;&gt;server -->
<g id="diagram-11-edge4" class="edge">
<title>cli&#45;&gt;server</title>
<path fill="none" stroke="#78908f" stroke-width="1.8" d="M162.23,-34.81C199.63,-40.31 243.65,-46.79 279.35,-52.04"/>
<polygon fill="#78908f" stroke="#78908f" stroke-width="1.8" points="279.03,-54.65 286.83,-53.14 279.79,-49.46 279.03,-54.65"/>
<text text-anchor="middle" x="224.5" y="-52.8" font-family="Arial" font-size="14.00" fill="#536b72">cli</text>
</g>
</g>
</svg>
</div>
```
<!-- /inline-svg -->

::: {.takeaway}

One primary owner per behavior; lower layers test their own contracts.

:::

::: {.notes}

**Say:** The tests mirror architectural boundaries: client_server, server_relay, relay_worker, and cli. Public behavior belongs at the outermost boundary that can usefully observe it. Private-boundary cases cover their own protocol seams rather than replicating every public message at every layer.

**Show:** Reuse the process topology with test-boundary names on the arrows. This ties the development method to the earlier architecture rather than presenting a directory tree with no explanation.

**Sources:** [Boundary test guide](https://github.com/t-kalinowski/mcp-console/blob/main/tests/boundaries/README.md)

:::



## YAML is what the reviewer reads {#yaml-transcripts .deck-slide .snapshot-slide}
::: {.kicker}

Addendum · Development practices

:::

::::: {.two-col}

:::: {.left}

::: {.code-panel}

::: {.panel-label}

Checked-in YAML · request

:::

```yaml
send:
  r: |
    cores <- parallel::detectCores()
    stopifnot(
      length(cores) == 1L,
      !is.na(cores),
      cores >= 1L
    )
    writeLines("R core detection available")
```

:::

::::

:::: {.right}

::: {.result-panel}

::: {.panel-label}

Checked-in YAML · result

:::

```yaml
result:
  content:
    - type: text
      text: |
        R core detection available
  isError: false
```

:::

::::

:::::

::: {.takeaway}

Readable code, arguments, output, and errors. Not escaped JSON strings.

:::

::: {.small-note}

Source excerpt: `detects_cpu_cores.yaml` · initialization document omitted.

:::

::: {.notes}

**Say:** The wire protocols are JSON-based, but the reviewable test transcripts are YAML. This is an actual checked-in snapshot excerpt: the test executes R, asserts that CPU detection returns a valid result, and prints a stable message. Humans can review the code and result without reading escaped JSON strings or snapshotting a machine-specific core count. YAML is the human review surface, not the production transport.

**Show:** Display the request and result from `tests/snapshots/client_server/r/test_runtime/detects_cpu_cores.yaml` side by side. The initial shared-handshake document is omitted, but the displayed request and result content come from the fetched source. This fixture was not rerun in the slide-build environment; do not present it as a new test execution.

**Sources:** [Boundary test guide](https://github.com/t-kalinowski/mcp-console/blob/main/tests/boundaries/README.md) · [Checked-in CPU detection snapshot](https://github.com/t-kalinowski/mcp-console/blob/main/tests/snapshots/client_server/r/test_runtime/detects_cpu_cores.yaml)

:::


## Normalize noise, not behavior {#test-normalization .deck-slide}
::: {.kicker}

Addendum · Development practices

:::

::::: {.two-col}

:::: {.left}

::: {.code-panel}

::: {.panel-label}

Unstable details

:::

```text
/tmp/run-8fd31/work.csv
worker pid: 27186
[waiting for stdin]
```

:::

::::

:::: {.right}

::: {.code-panel}

::: {.panel-label}

Schematic normalization

:::

```text
<WORKSPACE>/work.csv
worker pid: <PID>
[waiting for stdin]
```

:::

::::

:::::

::: {.takeaway}

Keep meaningful wording, ordering, and failure distinctions visible.

:::

::: {.notes}

**Say:** Temporary paths, process identities, and similar unstable details should not obscure behavioral review. Normalize explicitly and narrowly. Preserve the fields, output, ordering, and failure distinctions the contract is meant to protect. The actual project normalizers and snapshot metadata are more specific than this schematic example.

**Show:** Use before/after panels with only incidental details changed. Call out that normalization is not a license to delete inconvenient evidence from a test.

**Sources:** [Boundary test guide](https://github.com/t-kalinowski/mcp-console/blob/main/tests/boundaries/README.md)

:::



## Synchronize; do not sleep and hope {#test-determinism .deck-slide}
::: {.kicker}

Addendum · Development practices

:::

::: {.big-statement}

Start → checkpoint → release → assert

:::

::: {.takeaway}

Explicit gates for output, cancellation, restart, and cleanup.

:::

::: {.notes}

**Say:** Concurrency and liveness tests need causal synchronization. Fixtures use gates and checkpoints so the test knows when the relevant state has actually been reached. Arbitrary sleeps are not proof that output was drained, an interrupt was delivered, or an owned resource was retired.

**Show:** Show one deterministic sequence and connect it to the timing and shutdown contracts from the talk. This is a development-practice slide, not an implementation recipe for every fixture.

**Sources:** [Boundary test guide](https://github.com/t-kalinowski/mcp-console/blob/main/tests/boundaries/README.md)

:::



## Portable behavior; real capability checks {#test-capabilities .deck-slide}
::: {.kicker}

Addendum · Development practices

:::

::::: {.two-col}

:::: {.left}

::: {.big-statement}

Shared cases<br>direct + sandboxed

:::

::::

:::: {.right}

::: {.big-statement}

Capability fixtures<br>SSH · Docker · SBX

:::

::::

:::::

::: {.takeaway}

Fake peers establish protocol contracts; real targets establish target behavior.

:::

::: {.notes}

**Say:** Reuse portable cases across execution modes and centralize capability discovery. A skipped target fixture is not target validation. Deterministic peers can establish orchestration behavior, but they cannot establish real container or microVM cleanup. Keep real-target evidence distinct from simulated protocol coverage.

**Show:** Use the two testing levels rather than an all-green platform matrix. Include lifecycle and security assertions where a textual snapshot alone cannot represent the guarantee.

**Sources:** [Boundary test guide](https://github.com/t-kalinowski/mcp-console/blob/main/tests/boundaries/README.md)

:::



## Snapshots are generated evidence {#snapshot-review .deck-slide}
::: {.kicker}

Addendum · Development practices

:::

::: {.code-panel}

::: {.panel-label}

Run, regenerate, review

:::

```bash
scripts/test --update BOUNDARY/SUITE::CASE

git diff -- tests/snapshots/
```

:::



::: {.notes}

**Say:** Generate transcripts from running tests, then review the resulting diff. Do not hand-edit the expectation to make a test pass. The value of readable YAML is that review can focus on the actual changed behavior, while assertions still enforce facts that a snapshot cannot show.

**Show:** Show the update command and a diff review, not a large wall of green tests. The selector is schematic; replace BOUNDARY/SUITE::CASE with a real selected fixture for a live development demonstration.

**Sources:** [Boundary test guide](https://github.com/t-kalinowski/mcp-console/blob/main/tests/boundaries/README.md)

:::



## Test the installed product {#installed-product-tests .deck-slide}
::: {.kicker}

Addendum · Development practices

:::

::::: {.two-col}

:::: {.left}

::: {.big-statement}

Source build<br>and wheels

:::

::::

:::: {.right}

::: {.big-statement}

R / Python clients<br>and native bundle

:::

::::

:::::

::: {.takeaway}

Packaging, launch, schemas, images, shutdown, and cleanup are part of the contract.

:::

::: {.notes}

**Say:** The delivered product includes the executable, companion binaries, Python interfaces, the R wrapper, and launch/lifetime behavior. Running a development binary alone does not establish that the installed bundle or each adapter works. Integration examples and packaging checks should be part of the release evidence, with actual target validation reported separately.

**Show:** Finish the addendum by connecting the install story to the test strategy. Avoid claiming every proposed integration test already exists: distinguish observed project coverage from the release checklist in the companion author notes.

**Sources:** [Boundary test guide](https://github.com/t-kalinowski/mcp-console/blob/main/tests/boundaries/README.md) · [Release and installed bundle](https://github.com/t-kalinowski/mcp-console/blob/main/RELEASE.md) · [Python clients and integrations](https://github.com/t-kalinowski/mcp-console/blob/main/docs/PYTHON.md) · [R package interface](https://github.com/t-kalinowski/mcp-console/blob/main/r/README.md)

:::
